---
title: "SMART-MS Figure 5 – CSF proteomics analysis"
author: "Dimitrios Kleftogiannis"
output:
  html_document:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---
  
### Contact
  
Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.

You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2026, University of Bergen (UiB) and Neuro-SysMed, Norway

### Initiation and parameters setup - MAYBE EDITS ARE NEEDED FOR DIRS 
```{r}

suppressPackageStartupMessages({
library(RColorBrewer)
library(gplots) 
library(ggplot2)
library(gdata)
library(gridExtra)
library(grid)
library(reshape2)
library(ggthemes)
library(ggpubr)
library(dplyr)
library(umap)
library(ComplexHeatmap)
library(ConsensusClusterPlus)
library(readxl)
library(igraph)
library(matrixStats)
library(ggforce)
library(ggraph)
library(ggridges)
library(tidyr)  
library(ggrepel)
library(broom)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(writexl)
library(forcats)
library(clusterProfiler)
library(enrichplot)
library(patchwork)
library(stringr)
  
})

#EDIT HERE IF NEEDED - SET YOUR WORKING DIR APPROPRIATELY
source("R/smartms_config.R")
source("R/io_zenodo.R")
#this is more elegant than before
cfg <- get_smartms_config()
# -----------------------------
# Parameters (EDIT THESE)
# -----------------------------
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
                   "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
                   "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
                   "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
                   "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names.info <- c('type','type','type','type','type','state','type','type',
                        'state','type','state','state','type','state','state','state',
                        'state','state','state','state','state','state','type','state',
                        'state','type','type','state','state','type','state','type',
                        'state','type','state','type','state','type','state','state','type')

params <- list(
  # Direct download URL to Zenodo file that contains allCells_combined.RDa
  # (Use the "Download" link for the file on Zenodo)
  zenodo_url = "PASTE_ZENODO_DIRECT_DOWNLOAD_URL_HERE",

  #EDIT HERE IF NEEDED
  #data_path = "PASTE_LOCAL_DOWNLOAD_FILE_HERE",
  data_path = paste('/Home/siv32/dkl081/Desktop/SMART/SMART_RAMMS_harmonised/Proteomics data analysis Normalized data.xlsx',sep =''),
  
  #EDIT HERE IF NEEDED
  out_base = "outputs/SMARTMS_publication",

  # Baseline definition for SMART-MS in your original notebook
  cohort = "SMART",
  timepoints = c("V4","V8","V10"),
  baseline_timePoint = "V4",
  baseline_label = "BL"

)

# -----------------------------
# Helpers
# -----------------------------
ensure_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

download_if_missing <- function(url, dest_path) {
  if (file.exists(dest_path)) return(invisible(dest_path))
  ensure_dir(dirname(dest_path))
  message("Downloading dataset from Zenodo...")
  utils::download.file(url, destfile = dest_path, mode = "wb", quiet = FALSE)
  message("Saved to: ", dest_path)
  invisible(dest_path)
}

# Output directories for this figure
dirs <- list(
  figures = file.path(params$out_base, "figures", "fig5"),
  tables  = file.path(params$out_base, "tables",  "fig5"),
  qc      = file.path(params$out_base, "qc",      "fig5")
)
lapply(dirs, ensure_dir)

# Colors for time points (as in your original notebook)
color_TimePoints <- c("deepskyblue2", "magenta2", "indianred2")
names(color_TimePoints) <- params$timepoints

set.seed(params$seed)
```


### Load cleaned proteomics data (Zenodo → local)
```{r}
download_if_missing(params$zenodo_url, params$data_path)
df <- readxl::read_excel(params$data_path, sheet = 1)
stopifnot(exists("df"))

#manipulate the data
colRenaming <- c('S1','S1',
                 'S4','S4',
                 'S7','S7',
                 'S10','S10',
                 'S13','S13',
                 'S16','S16',
                 'S2','S2',
                 'S5','S5',
                 'S8','S8',
                 'S11','S11',
                 'S14','S14',
                 'S17','S17',
                 'S3','S3',
                 'S6','S6',
                 'S9','S9',
                 'S12','S12',
                 'S15','S15',
                 'S18','S18')

colTimePoint <- c('V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8',
                  'V4','V8')

a <- as.data.frame(t(df))
dt <- data.frame(Patient=colRenaming,
                 timePoint=colTimePoint)
proteinData <- cbind(dt,a[3:38,])
colnames(proteinData)[3:1327] <- df$Accession
proiteinFullNames <- df$Description

```


### Clinical design metadata 
Participant-level clinical covariates are not distributed in this public repository.
The full analysis requires a restricted local file with approved access.
Please request the clinical_covariates_SMART.RData file

```{r}

############################### trial info about time points ############################### 
# Participant-level clinical covariates are not distributed in this public repository.
# The full analysis requires a restricted local file with approved access.
# Please place the file at:
clinical_file <- 'clinical_covariates_SMART.RData'

if (!file.exists(clinical_file)) {
  stop(
    paste(
      "Restricted clinical covariate file not found:",
      clinical_file,
      "\nThis public repository does not include participant-level demographic metadata.",
      "\nApproved users should obtain the file from the study authors/data custodian",
      "and place it in the path above before running the full analysis."
    )
  )
}

load('clinical_covariates_SMART.RData')


```


### Dimensionality reduction and visualisation of the full proteome
```{r PCA,cache=TRUE,echo=FALSE}

# --- 1A) Make one metadata table with both timepoints
meta_V4 <- df_trial_V4 %>%
  mutate(timePoint = "V4")

meta_V8 <- df_trial_V8 %>%
  mutate(timePoint = "V8")

meta_all <- bind_rows(meta_V4, meta_V8) %>%
  # keep only the columns we need (add more if you want)
  select(Patient, timePoint, Treatment, Sex, Period, Age, Sequence)

# --- 1B) Attach metadata to proteinData
# proteinData has Patient + timePoint, then proteins
dat <- proteinData %>%
  left_join(meta_all, by = c("Patient", "timePoint"))

# Identify protein columns = everything AFTER Patient + timePoint
# (We will exclude the clinical columns we added via join)
protein_cols <- setdiff(
  colnames(dat),
  c("Patient","timePoint","Treatment","Sex","Period","Age","Sequence")
)

X <- dat %>%
  select(all_of(protein_cols)) %>%
  as.matrix()
# rownames for nice plotting/tracking
rownames(X) <- paste(dat$Patient, dat$timePoint, sep = "_")

# Minimal NA handling for PCA/UMAP:
# Replace NA with protein-wise median (simple + transparent for landscape plots)
for (j in seq_len(ncol(X))) {
  nas <- is.na(X[, j])
  if (any(nas)) X[nas, j] <- median(X[, j], na.rm = TRUE)
}

#weird R convertions
X_num <- apply(X, 2, function(col) as.numeric(gsub(",", ".", col)))
X_num <- as.matrix(X_num)
X <- X_num


########################################################## PCA

pca <- prcomp(X, center = TRUE, scale. = TRUE)
library(tibble)
pca_df <- as.data.frame(pca$x[, 1:2]) %>%
  rownames_to_column("SampleID") %>%
  mutate(
    Patient   = dat$Patient,
    timePoint = dat$timePoint,
    Treatment = dat$Treatment,
    Sex       = dat$Sex,
    Age       = dat$Age
  )

var1 <- round(100 * (pca$sdev[1]^2 / sum(pca$sdev^2)), 1)
var2 <- round(100 * (pca$sdev[2]^2 / sum(pca$sdev^2)), 1)

pca_df[pca_df$timePoint=='V4','timePoint'] <- 'BL'
pca_df[pca_df$timePoint=='V8','timePoint'] <- '6m'
pca_df$timePoint <- factor(pca_df$timePoint,levels = c('BL','6m'))

o <- ggplot(pca_df, aes(PC1, PC2, color = Treatment, shape = timePoint,group=Patient)) +
  geom_point(size = 3.2, alpha = 0.9) +
  geom_line(color = 'black', alpha = 0.25,linewidth=0.2) +
  # Patient ID labels (repelled so they don't overlap)
  geom_text_repel(
    aes(label = Patient),
    size = 3.2,
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Baseline" = "grey70",
    "Control"  = "royalblue3",
    "MSC"      = "khaki3"
  )) +
  labs(
    title = "CSF proteomics landscape - (PCA, Baseline → 6m)",
    x = paste0("PC1 (", var1, "%)"),
    y = paste0("PC2 (", var2, "%)"),
    color = "Treatment",
    shape = "Time point"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

ggsave(file.path(dirs$figures, "Figure_5_PCA.pdf"), o, width = 12, height = 9)

####################################################### UMAP

set.seed(1)
um <- umap(X, random_state=123, verbose =T)
umap_df <- as.data.frame(um$layout) %>%
  setNames(c("UMAP1","UMAP2")) %>%
  mutate(
    SampleID  = rownames(X),
    Patient   = dat$Patient,
    timePoint = dat$timePoint,
    Treatment = dat$Treatment,
    Sex       = dat$Sex,
    Age       = dat$Age
  )

umap_df[umap_df$timePoint=='V4','timePoint'] <- 'BL'
umap_df[umap_df$timePoint=='V8','timePoint'] <- '6m'
umap_df$timePoint <- factor(umap_df$timePoint,levels = c('BL','6m'))

o <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Treatment, shape = timePoint,group=Patient)) +
  geom_point(size = 3.2, alpha = 0.9) +
  geom_line(color = 'black', alpha = 0.25,linewidth=0.2) +
  # Patient ID labels (repelled so they don't overlap)
  geom_text_repel(
    aes(label = Patient),
    size = 3.2,
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Baseline" = "grey70",
    "Control"  = "royalblue3",
    "MSC"      = "khaki3"
  )) +
  labs(
    title = "CSF proteomics landscape - (UMAP, Baseline → 6m)",
    color = "Treatment",
    shape = "Time point"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )


ggsave(file.path(dirs$figures, "Figure_5_UMAP.pdf"), o, width = 12, height = 9)

########################################################## MDS
# Euclidean distance between samples
d <- dist(X)

# Classical MDS
mds <- cmdscale(d, k = 2)

mds_df <- as.data.frame(mds) %>%
  setNames(c("MDS1", "MDS2")) %>%
  rownames_to_column("SampleID") %>%
  mutate(
    Patient   = dat$Patient,
    timePoint = dat$timePoint,
    Treatment = dat$Treatment,
    Sex       = dat$Sex,
    Age       = dat$Age
  )

mds_df[mds_df$timePoint=='V4','timePoint'] <- 'BL'
mds_df[mds_df$timePoint=='V8','timePoint'] <- '6m'
mds_df$timePoint <- factor(mds_df$timePoint,levels = c('BL','6m'))

o <- ggplot(mds_df, aes(MDS1, MDS2, color = Treatment, shape = timePoint,group=Patient)) +
  geom_point(size = 3.2, alpha = 0.9) +
  geom_line(color = 'black', alpha = 0.25,linewidth=0.2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.3) +
  # Patient ID labels (repelled so they don't overlap)
  geom_text_repel(
    aes(label = Patient),
    size = 3.2,
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Baseline" = "grey70",
    "Control"  = "royalblue3",
    "MSC"      = "khaki3"
  )) +
  labs(
    title = "CSF proteomics landscape - (MDS, Baseline → 6m)",
    color = "Treatment",
    shape = "Time point"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

ggsave(file.path(dirs$figures, "Figure_5_MDS.pdf"), o, width = 12, height = 9)

```

### Modelling - Compute Deltas V8-V4 and perform lm fit
```{r differential modelling, echo=FALSE, cache=TRUE,eval=TRUE}

proteinData <- data.frame(Patient=dat$Patient,
                         timePoint=dat$timePoint,
                         Treatment=dat$Treatment,
                         Sex=dat$Sex,
                         Period=dat$Period,
                         Age=dat$Age,
                         Sequence=dat$Sequence)

proteinDat <- cbind(proteinData,X)

proteinDat %>% dplyr::count(timePoint)
proteinDat %>% distinct(Patient) %>% nrow()
proteinDat %>% dplyr::count(Patient) %>% dplyr::count(n)

#1
meta_cols <- c("Patient","timePoint","Treatment","Sex","Period","Age","Sequence")
protein_cols <- setdiff(colnames(proteinDat), meta_cols)

#build table
# Split baseline and 6m
dat_V4 <- proteinDat %>% filter(timePoint == "V4")
dat_V8 <- proteinDat %>% filter(timePoint == "V8")

# Ensure same patient order
dat_V4 <- dat_V4 %>% arrange(Patient)
dat_V8 <- dat_V8 %>% arrange(Patient)
stopifnot(all(dat_V4$Patient == dat_V8$Patient))

#compute delta
delta_mat <- dat_V8[, protein_cols] - dat_V4[, protein_cols]

#assembe modelling table
delta_table <- bind_cols(
  dat_V8 %>% select(Patient, Treatment, Sex, Age),
  delta_mat
)

#sanity checks
dim(delta_table)     # should be 18 x (4 + 1325)
table(delta_table$Treatment)
summary(as.matrix(delta_table[, protein_cols]))

#distribution overview - just for sanity
hist(as.numeric(as.matrix(delta_table[, protein_cols])), 
     breaks = 100, main = "Distribution of Δ values")


###### modelling starts here
# Make sure types are correct
delta_table <- delta_table %>%
  mutate(
    Age_c = scale(Age, center = TRUE, scale = FALSE)[,1],
    Treatment = factor(Treatment, levels = c("Control","MSC")),
    Sex = factor(Sex)
  )

#modelling function
model_results <- lapply(protein_cols, function(p) {
  
  df <- delta_table %>%
    select(Treatment, Age_c, Sex, all_of(p)) %>%
    rename(y = all_of(p))
  
  fit <- lm(y ~ Treatment + Age_c + Sex, data = df)
  
  broom::tidy(fit) %>%
    filter(term == "TreatmentMSC") %>%
    mutate(Protein = p)
})

results <- bind_rows(model_results)

results <- results %>%
  mutate(FDR = p.adjust(p.value, method = "BH"))


results <- results %>%
  mutate(
    direction = ifelse(estimate > 0, "Up in MSC", "Down in MSC"),
    logP = -log10(p.value),
    logFDR = -log10(FDR)
  )


sig_005 <- results %>% filter(FDR < 0.05)
sig_01  <- results %>% filter(FDR < 0.1)

ranked_list <- results %>%
  arrange(desc(estimate)) %>%
  select(Protein, estimate)

#sanity checks
summary(results$estimate)
summary(results$p.value)
sum(results$FDR < 0.05)

head(results[order(results$p.value), ], 10)

###########################################################################################################
#Volcano with annotation
###########################################################################################################
# Bioconductor annotation packages
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("AnnotationDbi", quietly = TRUE)) BiocManager::install("AnnotationDbi")
if (!requireNamespace("org.Hs.eg.db", quietly = TRUE)) BiocManager::install("org.Hs.eg.db")

## ---------------------------
## 1) Build volcano_df (with significance class)
## ---------------------------
volcano_df <- results %>%
  mutate(
    sig = case_when(
      FDR < 0.05 ~ "FDR < 0.05",
      FDR < 0.10 ~ "FDR < 0.10",
      TRUE ~ "NS"
    ),
    # remove isoform suffix if present, e.g. P12345-2 -> P12345
    Protein_clean = sub("-\\d+$", "", Protein)
  )

## ---------------------------
## 2) Map UniProt -> Gene symbol using org.Hs.eg.db (offline)
## ---------------------------
ids_clean <- unique(volcano_df$Protein_clean)
ids_clean <- ids_clean[!is.na(ids_clean) & ids_clean != ""]

mapping <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = ids_clean,
  keytype = "UNIPROT",
  columns = c("SYMBOL")
)

# Keep one symbol per UniProt (if multiple appear, keep the first)
mapping_one <- mapping %>%
  filter(!is.na(SYMBOL), SYMBOL != "") %>%
  distinct(UNIPROT, .keep_all = TRUE)

## ---------------------------
## 3) Attach gene symbols + create final label
## ---------------------------
volcano_df <- volcano_df %>%
  left_join(mapping_one, by = c("Protein_clean" = "UNIPROT")) %>%
  mutate(
    label = ifelse(is.na(SYMBOL) | SYMBOL == "", Protein, SYMBOL)
  )

## ---------------------------
## 4) (Optional) quick mapping sanity checks
## ---------------------------
cat("Mapped fraction:",
    mean(!(is.na(volcano_df$SYMBOL) | volcano_df$SYMBOL == "")),
    "\n")

# Show some examples
#print(volcano_df %>% select(Protein, Protein_clean, SYMBOL, label) %>% distinct() %>% head(20))
# set your effect threshold for "big effect"
effect_cut <- 1

## LIST OF KNOWN CONTAMINATION GENES
contam_genes <- c(
  "HBA1","HBB","VIM","TUBB2A","TUBA1A","HSPB1",
  "KRT1","KRT9","KPRP","PLEC","DSG1","DSP","AHNAK",
  "JUP","EEF2","CA3","DCD"
)

volcano_df2 <- volcano_df %>%
  mutate(
    is_big = abs(estimate) >= effect_cut,
    is_contam = label %in% contam_genes,  # uses gene symbols (your 'label' column)
    class = case_when(
      FDR < 0.05 ~ "FDR < 0.05",
      FDR < 0.10 ~ "FDR < 0.10",
      #is_big & is_contam ~ "Contamination/housekeeping",
      #is_big ~ paste0("High effect (|β| ≥ ", effect_cut, ")"),
      TRUE ~ "NS"
    )
  )

label_df <- volcano_df2 %>%
  filter(FDR < 0.10 | (abs(estimate) >= effect_cut & is_contam))

write_xlsx(label_df, file.path(dirs$tables, "Volcano_labeled_proteins.xlsx"))

o <- ggplot(volcano_df2, aes(estimate, logP)) +
  geom_point(aes(color = class), alpha = 0.85, size = 2.8) +
  
  # Thresholds
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",color='red3') +
  geom_hline(yintercept = -log10(0.10), linetype = "dotted",color='red3') +
  geom_vline(xintercept = c(-effect_cut, effect_cut), linetype = "dashed",color='red3') +
  
  # Labels
  geom_text_repel(
    data = label_df,
    aes(label = label),
    size = 4,
    max.overlaps = Inf
  ) +
  
  scale_color_manual(values = c(
    "FDR < 0.05" = "#D62728",
    "FDR < 0.10" = "#FF7F0E",
    "Contamination/housekeeping" = "#1F77B4",
    "NS" = "grey75")) +
  labs(
    x = "MSC effect on Delta protein (log2 scale)",
    y = expression(-log[10](p)),
    title = "Volcano plot — MSC effect on CSF proteome",
    color = "Category"
  ) +
  theme_classic() +
  theme(legend.position = "bottom")


ggsave(file.path(dirs$figures, "Figure_5_Volcano.pdf"), o, width = 12, height = 9)

```

### Visualisation of patients stratified by clinical reactivity
```{r combined stratification, echo=FALSE, cache=TRUE,eval=TRUE}

#clinical adverse events and MRI findings
MSC_reactivity <- data.frame(Patient=cfg$patientLevels_SMART,
                             Fever_Pain=c(NA,
                                          'No',
                                          'Yes',
                                          'Yes',
                                          'No',
                                          'No',
                                          'No',
                                          'Yes',
                                          'No',
                                          'No',
                                          'No',
                                          'Yes',
                                          'No',
                                          'No',
                                          'Yes',
                                          'No',
                                          'Yes',
                                          'No'),
                             MRI_events=c(NA,
                                          'No',
                                          'Yes',
                                          'Yes',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'No',
                                          'Yes',
                                          'No',
                                          'No',
                                          'No')
                             )



# Merge adverse events into delta_table
delta_annot <- delta_table %>%
  left_join(MSC_reactivity %>% dplyr::select(Patient, Fever_Pain,MRI_events), by = "Patient") %>%
  mutate(
    Group = case_when(
      Treatment == "MSC" & Fever_Pain == "Yes" &   MRI_events == 'No'  ~ "MSC + AE no MRI",
      Treatment == "MSC" & Fever_Pain == "No"  &   MRI_events == 'No'  ~ "MSC no AE no MRI",
      Treatment == "MSC" & Fever_Pain == 'Yes' &   MRI_events == 'Yes' ~ "MSC + AE + MRI",
      Treatment == "Control" ~ "Control",
      
    ),
    Group = factor(Group, levels = c("Control",'MSC no AE no MRI',"MSC + AE no MRI",'MSC + AE + MRI'))
  )

#remove patient S1
delta_annot <- delta_annot[-1,]

#pick the proteins of interest
hits_01 <- volcano_df %>%
  filter(FDR < 0.10) %>%
  distinct(Protein_clean, SYMBOL, FDR, estimate, sig)
hits <- hits_01

plot_df <- delta_annot %>%
  dplyr::select(Patient, Treatment, Sex, Age, Group, all_of(hits$Protein_clean)) %>%
  pivot_longer(
    cols = all_of(hits$Protein_clean),
    names_to = "Protein_clean",
    values_to = "Delta"
  ) %>%
  left_join(hits %>% dplyr::select(Protein_clean, SYMBOL, FDR), by = "Protein_clean") %>%
  mutate(
    SYMBOL = ifelse(is.na(SYMBOL) | SYMBOL == "", Protein_clean, SYMBOL),
    SYMBOL = fct_reorder(SYMBOL, Delta, .fun = median, .na_rm = TRUE)
  )

#5) Targeted visualisation #2: Rank proteins by volcano effect, then facet
order_vec <- hits %>% arrange(FDR) %>% pull(SYMBOL)

plot_df2 <- plot_df %>%
  mutate(SYMBOL = factor(as.character(SYMBOL), levels = order_vec))


# --- Define colors ---
group_cols <- c(
  "Control"   = "royalblue3",
  "MSC no AE no MRI" = "khaki",
  "MSC + AE no MRI"  = "khaki3",
  "MSC + AE + MRI"  = "khaki4"
)

ae_cols <- c("No" = "grey25", "Yes" = "firebrick3")
p_delta_ordered <- ggplot(plot_df2, aes(x = Group, y = Delta)) +
      geom_hline(yintercept = 0, linetype = "dashed", color='red',alpha = 0.5) +
      
      # Boxplots colored by Group
      geom_boxplot(aes(fill = Group), 
                   outlier.shape = NA, 
                   width = 0.38, 
                   alpha = 0.48,
                   notch = FALSE) +
      
      # Points: AE highlighted by color
      geom_point(
        aes(color = 'gray88'),
        size = 1.18, alpha = 0.68,
        position = position_jitter(width = 0.12, height = 0)
      ) +
      geom_text(
        data = subset(plot_df2, Treatment != "Control"),
        aes(label = Patient),
        size = 1.4,
        vjust = -0.5
      )+

      facet_wrap(~ SYMBOL, scales = "free_y", ncol = 4) +
      scale_fill_manual(values = group_cols) +
      scale_color_manual(values = ae_cols) +
      labs(
        x = NULL,
        y = expression(Delta~"(6m - BL)"),
        title = "Differentially enriched CSF proteins",
        #subtitle = "Boxplots by group; patients highlighted and labeled",
        fill = "Group",
        color = "Reactivity"
      ) +
      theme_classic() +
      theme(
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_rect(fill = "grey95", color = NA),
        strip.text = element_text(face = "bold"),
        aspect.ratio = 0.8
  )

ggsave(file.path(dirs$figures, "Expression_diffGenes.pdf"), p_delta_ordered, width = 12, height = 9)

################################################################################
#contamination 
################################################################################

contam_genes <- c(
  "HBA1","KPRP","KRT9","KRT1","PLEC",
  "HBB","VIM","DCD","DSP")

contam_hits <- volcano_df %>%
  filter(SYMBOL %in% contam_genes) %>%
  distinct(Protein_clean, SYMBOL) %>%
  arrange(match(SYMBOL, contam_genes))

# keep only those that exist in delta_table
contam_hits <- contam_hits %>%
  filter(Protein_clean %in% colnames(delta_table))

contam_plot_df <- delta_annot %>%
  dplyr::select(Patient, Treatment, Sex, Age, Group, all_of(contam_hits$Protein_clean)) %>%
  pivot_longer(
    cols = all_of(contam_hits$Protein_clean),
    names_to = "Protein_clean",
    values_to = "Delta"
  ) %>%
  left_join(contam_hits, by = "Protein_clean") %>%
  mutate(
    SYMBOL = factor(SYMBOL, levels = contam_genes[contam_genes %in% unique(SYMBOL)])
  )

p_contam <- ggplot(contam_plot_df, aes(x = Group, y = Delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", color='red',alpha = 0.5) +
  
  # Boxplots colored by Group
  geom_boxplot(aes(fill = Group), 
               outlier.shape = NA, 
               width = 0.38, 
               alpha = 0.48,
               notch = FALSE) +
  
  # Points: AE highlighted by color
  geom_point(
    aes(color = 'gray88'),
    size = 1.18, alpha = 0.68,
    position = position_jitter(width = 0.12, height = 0)
  ) +
  geom_text(
    data = subset(contam_plot_df, Treatment != "Control"),
    aes(label = Patient),
    size = 1.4,
    vjust = -0.5
  )+
  
  facet_wrap(~ SYMBOL, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = group_cols) +
  scale_color_manual(values = ae_cols) +
  labs(
    x = NULL,
    y = expression(Delta~"(6m - BL)"),
    title = "Contamination/high-effect markers: CSF protein changes by group",
    #subtitle = "Boxplots by group; patients highlighted and labeled",
    fill = "Group",
    color = "Reactivity"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(face = "bold"),
    aspect.ratio = 0.8
  )

ggsave(file.path(dirs$figures, "Expression_ContamGenes.pdf"), p_contam, width = 12, height = 9)

################################################################################
#inflamation
################################################################################
csf_inflam_genes <- c(
  "CRP", "SAA1", "SAA2", "HP", "ORM1", "ORM2",
  "FGA", "FGB", "FGG", "SERPINA1", "SERPINA3",
  "C3", "C4A", "C4B","IL6ST","LGALS3","A2M"
)

inflam_hits <- volcano_df %>%
  filter(SYMBOL %in% csf_inflam_genes) %>%
  distinct(Protein_clean, SYMBOL) %>%
  arrange(match(SYMBOL, csf_inflam_genes))

# keep only those that exist in delta_table
inflam_hits <- inflam_hits %>%
  filter(Protein_clean %in% colnames(delta_table))

inflam_plot_df <- delta_annot %>%
  dplyr::select(Patient, Treatment, Sex, Age, Group, all_of(inflam_hits$Protein_clean)) %>%
  pivot_longer(
    cols = all_of(inflam_hits$Protein_clean),
    names_to = "Protein_clean",
    values_to = "Delta"
  ) %>%
  left_join(inflam_hits, by = "Protein_clean") %>%
  mutate(
    SYMBOL = factor(SYMBOL, levels = csf_inflam_genes[csf_inflam_genes %in% unique(SYMBOL)])
  )

p_inflam <- ggplot(inflam_plot_df, aes(x = Group, y = Delta)) +
  geom_hline(yintercept = 0, linetype = "dashed", color='red',alpha = 0.5) +
  
  # Boxplots colored by Group
  geom_boxplot(aes(fill = Group), 
               outlier.shape = NA, 
               width = 0.38, 
               alpha = 0.48,
               notch = FALSE) +
  geom_point(
    aes(color = 'gray88'),
    size = 1.18, alpha = 0.68,
    position = position_jitter(width = 0.12, height = 0)
  ) +
  geom_text(
    data = subset(inflam_plot_df, Treatment != "Control"),
    aes(label = Patient),
    size = 1.4,
    vjust = -0.5
  )+
  facet_wrap(~ SYMBOL, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = group_cols) +
  scale_color_manual(values = ae_cols) +
  labs(
    x = NULL,
    y = expression(Delta~"(6m - BL)"),
    title = "Inflammation markers: CSF protein changes by group",
    #subtitle = "Boxplots by group; patients highlighted and labeled",
    fill = "Group",
    color = "Reactivity"
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey95", color = NA),
    strip.text = element_text(face = "bold"),
    aspect.ratio = 0.8
  )

ggsave(file.path(dirs$figures, "Expression_inflamGenes.pdf"), p_inflam, width = 12, height = 9)


```


### Pathway based analysis using GO
```{r GSEA, echo=FALSE, cache=TRUE,eval=TRUE}

ranked_list <- results %>%
  dplyr::select(Protein, estimate) %>%
  dplyr::arrange(desc(estimate))

# Convert UniProt → Gene Symbol (use your mapped volcano_df)
ranked_list <- ranked_list %>%
  dplyr::left_join(volcano_df %>% 
          dplyr::select(Protein, label), by = "Protein") %>%
  dplyr::distinct(label, .keep_all = TRUE) %>%
  dplyr::filter(!is.na(label))

# Named vector for GSEA
gene_ranks <- ranked_list$estimate
names(gene_ranks) <- ranked_list$label

# Sort decreasing
gene_ranks <- sort(gene_ranks, decreasing = TRUE)

gsea_go <- gseGO(
  geneList = gene_ranks,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "SYMBOL",
  minGSSize = 10,
  maxGSSize = 300,
  pvalueCutoff = 0.1,
  verbose = FALSE
)

#default visualisation
dotplot(gsea_go, showCategory = 15, title = "GSEA – GO Biological Process")
#dotplot(gsea_react, showCategory = 15, title = "GSEA – Reactome")
#BiocManager::install("enrichplot", ask = FALSE)

gseaplot2(gsea_go, geneSetID = 1, title = gsea_go@result$Description[1])
#examine the pathways
a <- head(gsea_go@result[, c("Description","NES","p.adjust")], 15)

################################# make the first plot with all proteins included
# ---- 0) Extract result table from gseGO object
dfp <- as.data.frame(gsea_go@result)
# ---- 1) Clean Description + compute leading-edge support
dfp <- dfp %>%
  mutate(
    Description = Description %>%
      str_replace_all("[\r\n]+", " ") %>%  # remove line breaks
      str_replace_all("\u2212|\u2010|\u2011|\u2012|\u2013|\u2014", "-") %>% # normalize unicode hyphens
      str_squish(),                        # trim + collapse spaces
    
    # Leading-edge genes are stored as "GENE1/GENE2/..."
    LeadingEdgeCount = ifelse(
      is.na(core_enrichment) | core_enrichment == "",
      0L,
      str_count(core_enrichment, "/") + 1L
    ),
    
    LeadingEdgeRatio = ifelse(setSize > 0, LeadingEdgeCount / setSize, NA_real_)
  ) %>%
  # drop any rows with missing descriptions just in case
  filter(!is.na(Description), Description != "")

# ---- 2) Select top 15 terms by pathway FDR
top_n <- 20
dfp_top <- dfp %>%
  arrange(p.adjust) %>%
  slice_head(n = top_n)

# ---- 3) Assign modules (your story grouping)
dfp_top <- dfp_top %>%
  mutate(
    Module = case_when(
      str_detect(str_to_lower(Description), "microtubule|cytoskeleton|tubulin|actin") ~ "Cytoskeletal remodeling",
      str_detect(str_to_lower(Description), "nucleobase|nucleotide|purine|phosphate|nucleoside") ~ "Nucleotide metabolism",
      str_detect(str_to_lower(Description), "carboxylic|oxoacid|organic acid|monocarboxylic") ~ "Organic acid metabolism",
      str_detect(str_to_lower(Description), "platelet|hemost|coagulat|acute-phase") ~ "Vascular / inflammatory interface",
      TRUE ~ "Other"
    )
  ) %>%
  # ---- 4) Order pathways nicely: first by Module, then by NES within Module
  group_by(Module) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  ungroup() %>%
  mutate(
    # order facets in a logical narrative order
    Module = factor(
      Module,
      levels = c("Vascular / inflammatory interface",
                 "Cytoskeletal remodeling",
                 "Nucleotide metabolism",
                 "Organic acid metabolism",
                 "Other")
    )
  )

dfp_top <- dfp_top %>% mutate(Description = str_wrap(as.character(Description), width = 35))

# Cap extremely small p.adjust to keep legend useful
dfp_top <- dfp_top %>%
  mutate(padj_cap = pmax(p.adjust, 1e-6))

o <- ggplot(dfp_top, aes(x = NES, y = reorder(Description, NES))) +
  geom_point(aes(size = LeadingEdgeCount, color = padj_cap), alpha = 0.9) +
  facet_grid(Module ~ ., scales = "free_y", space = "free_y", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_continuous(
    trans = "log10",
    breaks = c(1e-6, 1e-4, 1e-3, 1e-2, 5e-2, 1e-1),
    labels = scales::label_scientific(digits = 1),
    guide = guide_colorbar(
      barheight = unit(4, "mm"),
      barwidth  = unit(45, "mm"),
      title.position = "top"
    )
  ) +
  scale_size_continuous(
    range = c(2.2, 7.5),
    breaks = scales::pretty_breaks(n = 4),
    guide = guide_legend(
      title.position = "top",
      override.aes = list(alpha = 1)
    )
  ) +
  labs(
    title = "GO GSEA: MSC-associated CSF programs",
    x = "NES (Normalized Enrichment Score)",
    y = NULL,
    size = "Leading-edge count",
    color = "FDR (p.adjust)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    
    # Facet strips on the left, clean and readable
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, face = "bold", size = 11),
    strip.background = element_blank(),
    
    # Make it compact
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing.y = unit(0.25, "lines"),
    
    # Improve label readability
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 11),
    
    plot.title = element_text(face = "bold", size = 15),
    plot.margin = margin(8, 18, 8, 8)
  )


ggsave(file.path(dirs$figures, "Figure_5_GO_all_ordered.pdf"), o, width = 12, height = 9)

##########################################################################################
#remove known contaminants
##########################################################################################
contam_tier1 <- c("HBA1","HBB","KRT1","KRT9","KPRP","DSG1","DCD")
contam_tier2 <- c("PLEC","DSP","AHNAK","JUP","CA3")
contam_all <- unique(c(contam_tier1, contam_tier2))

gene_ranks_clean <- gene_ranks[ !names(gene_ranks) %in% contam_all ]

# sanity check
length(gene_ranks)        # before
length(gene_ranks_clean)  # after

gsea_go_clean <- gseGO(
  geneList = gene_ranks_clean,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "SYMBOL",
  minGSSize = 10,
  maxGSSize = 300,
  pvalueCutoff = 0.1,
  verbose = FALSE
)

####check the first 20 !!!
a <- head(gsea_go@result[, c("Description","NES","p.adjust")], 20)
b <- head(gsea_go_clean@result[, c("Description","NES","p.adjust")], 20)

p1 <- dotplot(gsea_go, showCategory = 20) + ggtitle("All proteins")
p2 <- dotplot(gsea_go_clean, showCategory = 20) + ggtitle("After contaminant removal")
#p1 + p2

####################################################################
# ---- 0) Extract result table from gseGO object
dfp <- as.data.frame(gsea_go_clean@result)

# ---- 1) Clean Description + compute leading-edge support
dfp <- dfp %>%
  mutate(
    Description = Description %>%
      str_replace_all("[\r\n]+", " ") %>%  # remove line breaks
      str_replace_all("\u2212|\u2010|\u2011|\u2012|\u2013|\u2014", "-") %>% # normalize unicode hyphens
      str_squish(),                        # trim + collapse spaces
    
    # Leading-edge genes are stored as "GENE1/GENE2/..."
    LeadingEdgeCount = ifelse(
      is.na(core_enrichment) | core_enrichment == "",
      0L,
      str_count(core_enrichment, "/") + 1L
    ),
    
    LeadingEdgeRatio = ifelse(setSize > 0, LeadingEdgeCount / setSize, NA_real_)
  ) %>%
  # drop any rows with missing descriptions just in case
  filter(!is.na(Description), Description != "")

# ---- 2) Select top 15 terms by pathway FDR
top_n <- 20
dfp_top <- dfp %>%
  arrange(p.adjust) %>%
  slice_head(n = top_n)

# ---- 3) Assign modules (your story grouping)
dfp_top <- dfp_top %>%
  mutate(
    desc_l = str_to_lower(Description),
    Module = case_when(
      # 1) Cytoskeletal remodeling / intracellular organization
      str_detect(desc_l,
                 "microtubule|cytoskeleton|tubulin|actin|protein localization to nucleus|nucleus|response to heat") ~
        "Cytoskeletal remodeling",

      # 2) Nucleotide / nucleic acid / biosynthetic programs
      str_detect(desc_l,
                 "nucleobase|nucleotide|purine|nucleoside|nucleic acid|organophosphate|phosphate|biosynthetic|macromolecule biosynthetic") ~
        "Nucleotide metabolism",

      # 3) Organic acid / central carbon metabolism
      str_detect(desc_l,
                 "carboxylic|oxoacid|organic acid|monocarboxylic") ~
        "Organic acid metabolism",

      # 4) Vascular / inflammatory interface
      str_detect(desc_l,
                 "platelet|hemost|coagulat|acute-phase|acute inflammatory|inflamm") ~
        "Vascular / inflammatory interface",

      TRUE ~ "Other"
    )
  )  %>%
  # ---- 4) Order pathways nicely: first by Module, then by NES within Module
  group_by(Module) %>%
  mutate(Description = fct_reorder(Description, NES)) %>%
  ungroup() %>%
  mutate(
    # order facets in a logical narrative order
    Module = factor(
      Module,
      levels = c("Vascular / inflammatory interface",
                 "Cytoskeletal remodeling",
                 "Nucleotide metabolism",
                 "Organic acid metabolism",
                 "Other")
    )
  )

dfp_top <- dfp_top %>% mutate(Description = str_wrap(as.character(Description), width = 35))

# Cap extremely small p.adjust to keep legend useful
dfp_top <- dfp_top %>%
  mutate(padj_cap = pmax(p.adjust, 1e-6))

o <- ggplot(dfp_top, aes(x = NES, y = reorder(Description, NES))) +
  geom_point(aes(size = LeadingEdgeCount, color = padj_cap), alpha = 0.9) +
  facet_grid(Module ~ ., scales = "free_y", space = "free_y", switch = "y") +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  scale_color_continuous(
    trans = "log10",
    breaks = c(1e-6, 1e-4, 1e-3, 1e-2, 5e-2, 1e-1),
    labels = scales::label_scientific(digits = 1),
    guide = guide_colorbar(
      barheight = unit(4, "mm"),
      barwidth  = unit(45, "mm"),
      title.position = "top"
    )
  ) +
  scale_size_continuous(
    range = c(2.2, 7.5),
    breaks = scales::pretty_breaks(n = 4),
    guide = guide_legend(
      title.position = "top",
      override.aes = list(alpha = 1)
    )
  ) +
  labs(
    title = "GO GSEA: MSC-associated CSF programs",
    x = "NES (Normalized Enrichment Score)",
    y = NULL,
    size = "Leading-edge count",
    color = "FDR (p.adjust)"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw(base_size = 13) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_text(size = 11),
    legend.text  = element_text(size = 10),
    
    # Facet strips on the left, clean and readable
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, face = "bold", size = 11),
    strip.background = element_blank(),
    
    # Make it compact
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.spacing.y = unit(0.25, "lines"),
    
    # Improve label readability
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(size = 10),
    
    plot.title = element_text(face = "bold", size = 15),
    plot.margin = margin(8, 18, 8, 8)
  )

ggsave(file.path(dirs$figures, "Figure_5_GO_filtered.pdf"), o, width = 12, height = 9)

#show some extra plots
#gseaplot2(gsea_go, geneSetID = 1, title = gsea_go@result$Description[1])
sel_sets <- c(
  "acute-phase response",
  "platelet aggregation",
  "cytoskeleton organization",
  "microtubule-based process",
  "nucleotide metabolic process"
)

p <- list()
myC <- 1
for (s in sel_sets) {
  i <- which(gsea_go@result$Description == s)
  p <- gseaplot2(gsea_go, geneSetID = i, title = s)
  
  myfile <- paste(file.path(dirs$figures,'Extra_gsea_p_'),i,'.pdf',sep ='')
  ggsave(myfile, p, width = 12, height = 9)
}

```


### Functional programs  and scores in CSF
```{r functional programs, echo=FALSE, cache=TRUE,eval=TRUE}

# 1) Identify protein columns (everything after metadata)
meta_cols <- c("Patient", "Treatment", "Sex", "Age",'Age_c')
prot_cols <- setdiff(colnames(delta_table), meta_cols)

# 2) Clean UniProt IDs (remove isoform suffix like P12345-2)
prot_cols_clean <- sub("-\\d+$", "", prot_cols)

# 3) Map UniProt -> Gene symbol
map <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = unique(prot_cols_clean),
  keytype = "UNIPROT",
  columns = c("SYMBOL")
) %>%
  filter(!is.na(SYMBOL), SYMBOL != "") %>%
  distinct(UNIPROT, .keep_all = TRUE)

# 4) Build a lookup for all your columns (including unmapped)
lookup <- tibble(
  Protein = prot_cols,
  UNIPROT = prot_cols_clean
) %>%
  left_join(map, by = "UNIPROT") %>%
  mutate(Gene = ifelse(is.na(SYMBOL) | SYMBOL == "", UNIPROT, SYMBOL))

# Optional: how many mapped?
cat("Mapped fraction:",
    mean(!is.na(lookup$SYMBOL) & lookup$SYMBOL != ""),
    "\n")

#Helper to compute a program score from Δ values
# Helper: score = mean(z-scored proteins) across a gene set
score_program_from_delta <- function(delta_df, lookup, gene_set, meta_cols = c("Patient","Treatment","Sex","Age")) {
  # Which protein columns correspond to these genes?
  cols <- lookup %>%
    filter(Gene %in% gene_set) %>%
    pull(Protein) %>%
    unique()

  cols <- intersect(cols, colnames(delta_df))
  if (length(cols) < 3) {
    warning("Gene set has <3 proteins present in delta_table after mapping. Returning NA.")
    out <- delta_df[, meta_cols, drop = FALSE]
    out$score <- NA_real_
    out$n_proteins <- length(cols)
    return(out)
  }

  mat <- as.matrix(delta_df[, cols, drop = FALSE])

  # Z-score each protein across patients (important so one protein doesn't dominate)
  mat_z <- scale(mat)

  score <- rowMeans(mat_z, na.rm = TRUE)

  out <- delta_df[, meta_cols, drop = FALSE]
  out$score <- score
  out$n_proteins <- length(cols)
  out
}

######################################## definitions of gene sets #################################
genes_interface_ecm <- c(
  "POSTN","SPARC","COMP",
  "COL1A1","COL1A2","COL3A1","COL5A1","COL11A1","COL6A6",
  "FN1","TNC","VCAN","LAMA2","LAMC1"
)

genes_innate_vascular <- c(
  # Complement
  "C3","C4A","C4B","C5","CFH","CFB",
  # Acute-phase / innate
  "CRP","SAA1","SAA2","HP","HPX","ORM1","ORM2","SERPINA1","SERPINA3","A2M","CP",
  # Coagulation / vascular tone
  "FGA","FGB","FGG"
)

genes_metcyto <- c(
  # Cytoskeleton / actin dynamics
  "CFL1","PFN1","ACTB","ACTG1",
  # Microtubules / transport
  "TUBA1A","TUBB2A","TUBB",
  # Stress / protein synthesis
  "HSPB1","EEF2"
  # Optional if present
  #"VIM"
)

#this is interesting to reflect tissue stress
genes_procedural_stress <- c(
  # Blood-derived
  "HBA1", "HBB",
  # Cytokeratins / epithelial
  "KRT1", "KPRP",
  # Junctional / epithelial scaffolds
  "DSG1", "DSP", "JUP",
  # Large structural scaffolds (released on damage)
  "AHNAK",
  # Muscle / erythrocyte-associated
  "CA3",
  # Skin / sweat associated
  "DCD",
  "VIM"
)

##################################################################################

csf_interface_score <- score_program_from_delta(delta_table, lookup, genes_interface_ecm)
csf_vascular_score   <- score_program_from_delta(delta_table, lookup, genes_innate_vascular)
csf_metcyto_score   <- score_program_from_delta(delta_table, lookup, genes_metcyto)
csf_stress_score   <- score_program_from_delta(delta_table, lookup, genes_procedural_stress)

# Rename score columns for clarity
csf_interface_score <- csf_interface_score %>% dplyr::rename(CSF_Interface = score, CSF_Interface_n = n_proteins)
csf_vascular_score   <- csf_vascular_score %>% dplyr::rename(CSF_Vascular = score, CSF_Vascular_n = n_proteins)
csf_metcyto_score   <- csf_metcyto_score %>% dplyr::rename(CSF_MetCytoskel = score, CSF_MetCytoskel_n = n_proteins)
csf_stress_score   <- csf_stress_score %>% dplyr::rename(CSF_ProcedStress = score, CSF_ProcedStress_n = n_proteins)


# Combine into one CSF program table
csf_programs <- csf_interface_score %>%
  left_join(csf_vascular_score %>% dplyr::select(Patient, CSF_Vascular, CSF_Vascular_n), by="Patient") %>%
  left_join(csf_metcyto_score %>% dplyr::select(Patient, CSF_MetCytoskel, CSF_MetCytoskel_n), by="Patient") %>%
  left_join(csf_stress_score %>% dplyr::select(Patient, CSF_ProcedStress, CSF_ProcedStress_n), by="Patient")

#remove patient S1
csf_programs <- csf_programs[-1,]


plot_df <- csf_programs %>%
  left_join(MSC_reactivity, by = "Patient") %>%
  mutate(
    Group = case_when(
      Treatment == "Control" ~ "Control",
      Treatment == "MSC" & Fever_Pain == "No" & MRI_events == "No" ~ "MSC no AE no MRI",
      Treatment == "MSC" & Fever_Pain == "Yes" & MRI_events == "No" ~ "MSC + AE no MRI",
      Treatment == "MSC" & Fever_Pain == "Yes" & MRI_events == "Yes" ~ "MSC + AE + MRI",
      TRUE ~ NA_character_
    ),
    Group = factor(
      Group,
      levels = c(
        "Control",
        "MSC no AE no MRI",
        "MSC + AE no MRI",
        "MSC + AE + MRI"
      )
    )
  )


plot_long <- plot_df %>%
  dplyr::select(Patient, Group,
         CSF_Interface, CSF_Vascular, CSF_MetCytoskel,CSF_ProcedStress) %>%
  tidyr::pivot_longer(
    cols = starts_with("CSF_"),
    names_to = "Program",
    values_to = "Score"
  ) %>%
  mutate(
    Program = recode(
      Program,
      CSF_Interface   = "CSF Interface/ECM",
      CSF_Vascular    = "CSF Innate/Vascular",
      CSF_MetCytoskel = "CSF Metabolic/Cytoskeletal",
      CSF_ProcedStress = "CSF Procedural Stress"
    )
  )

p_csf_programs <- ggplot(plot_long, aes(x = Group, y = Score)) +
      geom_hline(yintercept = 0, linetype = "dashed", color='red',alpha = 0.4) +
      geom_boxplot(
        aes(fill = Group),
        outlier.shape = NA,
        width = 0.45,
        alpha = 0.45
      ) +
      geom_point(
        size = 1.48,
        alpha = 0.68,
        position = position_jitter(width = 0.12)
      ) +
      geom_text(
        data = subset(plot_long, Group != "Control"),
        aes(label = Patient),
        size = 2.2,
        vjust = -0.5
      )+
      facet_wrap(~ Program, scales = "free_y", nrow = 2) +
      scale_fill_manual(
        values = c(
          "Control" = "royalblue3",
          "MSC no AE no MRI" = "khaki",
          "MSC + AE no MRI" = "khaki3",
          "MSC + AE + MRI" = "khaki4"
        )
      ) +
      labs(
        x = NULL,
        y = expression(Delta~"CSF program score (6m - BL)"),
        #title = "CSF functional program responses stratified by MSC reactivity",
        #subtitle = "Patients grouped by adverse events and MRI findings"
      ) +
      theme_classic(base_size = 12) +
      theme(
        legend.position = "none",
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 25, hjust = 1),
        aspect.ratio = 0.68
      )

ggsave(file.path(dirs$figures, "Figure_5_CSF_scores.pdf"), p_csf_programs, width = 12, height = 9)

###########################################################################################
#scatter plot of scores colored by group

o1 <- ggplot(plot_df,
       aes(x = CSF_Vascular, y = CSF_Interface, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, alpha = 0.9) +
  #geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  scale_color_manual(values = c(
    "Control" = "royalblue3",
    "MSC no AE no MRI" = "khaki",
    "MSC + AE no MRI" = "khaki3",
    "MSC + AE + MRI" = "khaki4"
  )) +
  xlim(-1.5,1.5)+
  ylim(-1.5,1.5)+
  theme_classic() +
  labs(
    x = "CSF vascular / inflammatory program",
    y = "CSF interface / ECM remodeling program",
    #title = "Decoupling of CSF interface remodeling from inflammatory stress",
    #subtitle = "MSC induces structural remodeling with variable vascular activation"
  )+
  theme(legend.position = 'bottom')

o2 <- ggplot(plot_df,
       aes(x = CSF_Vascular, y = CSF_MetCytoskel, , color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, alpha = 0.9) +
  #geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  scale_color_manual(values = c(
    "Control" = "royalblue3",
    "MSC no AE no MRI" = "khaki",
    "MSC + AE no MRI" = "khaki3",
    "MSC + AE + MRI" = "khaki4"
  )) +
  xlim(-1.5,1.5)+
  ylim(-1.5,1.5)+
  theme_classic() +
  labs(
    x = "CSF vascular / inflammatory program",
    y = "CSF Metabolic/Cytoskeletal",
    #title = "Decoupling of CSF Metabolic/Cytoskeletal remodelling from inflammatory stress",
    #subtitle = "MSC induces structural remodeling with variable vascular activation"
  )+
  theme(legend.position = 'bottom')

ggsave(file.path(dirs$figures, "CSF_scores_scatter1.pdf"), o1, width = 12, height = 9)
ggsave(file.path(dirs$figures, "CSF_scores_scatter2.pdf"), o2, width = 12, height = 9)

o3 <- ggplot(plot_df,
       aes(x = CSF_MetCytoskel, y = CSF_ProcedStress, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, alpha = 0.9) +
  #geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  scale_color_manual(values = c(
    "Control" = "royalblue3",
    "MSC no AE no MRI" = "khaki",
    "MSC + AE no MRI" = "khaki3",
    "MSC + AE + MRI" = "khaki4"
  )) +
  xlim(-1.5,1.5)+
  ylim(-1.5,1.5)+
  theme_classic() +
  labs(
    y = "CSF Procedural stress",
    x = "CSF Metabolic/Cytoskeletal",
    #title = "Decoupling of CSF Metabolic/Cytoskeletal remodelling from inflammatory stress",
    #subtitle = "MSC induces structural remodeling with variable vascular activation"
  )+
  theme(legend.position = 'bottom')

o4 <- ggplot(plot_df,
       aes(x = CSF_Interface, y = CSF_ProcedStress, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, alpha = 0.9) +
  #geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  scale_color_manual(values = c(
    "Control" = "royalblue3",
    "MSC no AE no MRI" = "khaki",
    "MSC + AE no MRI" = "khaki3",
    "MSC + AE + MRI" = "khaki4"
  )) +
  xlim(-1.5,1.5)+
  ylim(-1.5,1.5)+
  theme_classic() +
  labs(
    y = "CSF Procedural stress",
    x = "CSF interface / ECM remodeling program",
    #title = "Decoupling of CSF Metabolic/Cytoskeletal remodelling from inflammatory stress",
    #subtitle = "MSC induces structural remodeling with variable vascular activation"
  )+
  theme(legend.position = 'bottom')

o5 <- ggplot(plot_df,
       aes(x = CSF_Vascular, y = CSF_ProcedStress, color = Group)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.4) +
  geom_point(size = 3, alpha = 0.9) +
  #geom_smooth(method = "lm", se = FALSE, linetype = "dotted") +
  scale_color_manual(values = c(
    "Control" = "royalblue3",
    "MSC no AE no MRI" = "khaki",
    "MSC + AE no MRI" = "khaki3",
    "MSC + AE + MRI" = "khaki4"
  )) +
  xlim(-1.5,1.5)+
  ylim(-1.5,1.5)+
  theme_classic() +
  labs(
    y = "CSF Procedural stress",
    x = "CSF vascular / inflammatory program",
    #title = "Decoupling of CSF Metabolic/Cytoskeletal remodelling from inflammatory stress",
    #subtitle = "MSC induces structural remodeling with variable vascular activation"
  )+
  theme(legend.position = 'bottom')


ggsave(file.path(dirs$figures, "CSF_scores_scatter3.pdf"), o3, width = 12, height = 9)
ggsave(file.path(dirs$figures, "CSF_scores_scatter4.pdf"), o4, width = 12, height = 9)
ggsave(file.path(dirs$figures, "CSF_scores_scatter5.pdf"), o5, width = 12, height = 9)

#modelling and p values
summary(lm(formula = CSF_Interface ~ CSF_ProcedStress + Treatment, data = plot_df))
summary(lm(formula = CSF_MetCytoskel ~ CSF_ProcedStress + Treatment, data = plot_df))
summary(lm(formula = CSF_Vascular ~ CSF_ProcedStress + Treatment, data = plot_df))

```

