---
title: "SMART-MS Figure 3 – Longitudinal immune landscape"
author: "Dimitrios Kleftogiannis"
output:
  html_document:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---
  
### Contact
  
Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.

You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2026, University of Bergen (UiB) and Neuro-SysMed, Norway

### Initiation and parameters setup - MAYBE EDITS ARE NEEDED FOR DIRS 
```{r}

suppressPackageStartupMessages({
  library(ggplot2)
  library(umap)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(circlize)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(lme4)
  library(lmerTest)
  library(patchwork)
  library(ggrepel)
})

#EDIT HERE - SET YOUR WORKING DIR APPROPRIATELY
source("R/smartms_config.R")
source("R/io_zenodo.R")
#this is more elegant than before
cfg <- get_smartms_config()
# -----------------------------
# Parameters (EDIT THESE)
# -----------------------------
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
                   "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
                   "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
                   "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
                   "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names.info <- c('type','type','type','type','type','state','type','type',
                        'state','type','state','state','type','state','state','state',
                        'state','state','state','state','state','state','type','state',
                        'state','type','type','state','state','type','state','type',
                        'state','type','state','type','state','type','state','state','type')

params <- list(
  # Direct download URL to Zenodo file that contains allCells_combined.RDa
  # (Use the "Download" link for the file on Zenodo)
  zenodo_url = "PASTE_ZENODO_DIRECT_DOWNLOAD_URL_HERE",

  #EDIT HERE IF NEEDED
  data_path = "PASTE_LOCAL_DOWNLOAD_FILE_HERE",
  #example --> data_path = '/Home/siv32/dkl081/Desktop/SMART/SMART_RAMMS_harmonised/allCells_combined.RDa',
  
  #EDIT HERE IF NEEDED
  out_base = "outputs/SMARTMS_publication",

  # Baseline definition for SMART-MS data
  cohort = "SMART",
  timepoints = c("V4","V8","V10"),
  baseline_timePoint = "V4",
  baseline_label = "BL",

  # Reproducibility + runtime
  seed = 1,
  umap_max_cells = 50000,
  umap_markers = cfg$col.names, 
  
  # Modelling settings
  # cell type columns come from proportions (one per CellType)
  p_adjust_method = "BH",
  top_diag_n = 6  # number of top hits to show diagnostics
)

# -----------------------------
# Helpers
# -----------------------------
ensure_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

download_if_missing <- function(url, dest_path) {
  if (file.exists(dest_path)) return(invisible(dest_path))
  ensure_dir(dirname(dest_path))
  message("Downloading dataset from Zenodo...")
  utils::download.file(url, destfile = dest_path, mode = "wb", quiet = FALSE)
  message("Saved to: ", dest_path)
  invisible(dest_path)
}

# Output directories for this figure
dirs <- list(
  figures = file.path(params$out_base, "figures", "fig3_longitudinal"),
  tables  = file.path(params$out_base, "tables",  "fig3_longitudinal"),
  qc      = file.path(params$out_base, "qc",      "fig3_longitudinal")
)
lapply(dirs, ensure_dir)

# Colors for time points 
color_TimePoints <- c("deepskyblue2", "magenta2", "indianred2")
names(color_TimePoints) <- params$timepoints

set.seed(params$seed)
```

### Load cleaned CyTOF data (Zenodo → local)
```{r}
download_if_missing(params$zenodo_url, params$data_path)

# This should load an object named: allCells
load(params$data_path)
stopifnot(exists("allCells"))

# Robust naming
if (!("CellType" %in% colnames(allCells)) && ("cellType" %in% colnames(allCells))) {
  allCells$CellType <- allCells$cellType
}

# Required columns for Figure 3 + modelling
required_cols <- c("Cohort","timePoint","Patient","CellType")
missing_cols <- setdiff(required_cols, colnames(allCells))
if (length(missing_cols) > 0) stop("Missing required columns in allCells: ", paste(missing_cols, collapse=", "))

# Keep SMART only (you do SMART first in the original)
allCells <- allCells[allCells$Cohort == params$cohort, ]

# --- Subset by visit
tmp_V4  <- allCells[allCells$timePoint == "V4", ]
tmp_V8  <- allCells[allCells$timePoint == "V8", ]
tmp_V10 <- allCells[allCells$timePoint == "V10", ]

```


### Clinical design metadata 
```{r}

patientLevels_SMART <- cfg$patientLevels_SMART
################################################### V4
df_trial_V4 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline','Baseline',
                               'Baseline'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                       Period=c('0','0',
                               '0','0',
                               '0','0',
                               '0','0','0',
                               '0','0',
                               '0',
                               '0','0',
                               '0','0','0',
                               '0'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))
tmp_V4 <- merge(tmp_V4, df_trial_V4, by = "Patient", all.x = TRUE)

################################################### V8
df_trial_V8 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('Control','Control',
                               'MSC','MSC',
                               'Control','Control',
                               'MSC','MSC','MSC',
                               'Control','Control',
                               'MSC',
                               'Control','Control',
                               'MSC','MSC','MSC',
                               'Control'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                        Period=c('1','1',
                               '1','1',
                               '1','1',
                               '1','1','1',
                               '1','1',
                               '1',
                               '1','1',
                               '1','1','1',
                               '1'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))

tmp_V8 <- merge(tmp_V8, df_trial_V8, by = "Patient", all.x = TRUE)

################################################### V10
df_trial_V10 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('MSC','MSC',
                               'Control','Control',
                               'MSC','MSC',
                               'Control','Control','Control',
                               'MSC','MSC',
                               'Control',
                               'MSC','MSC',
                               'Control','Control','Control',
                               'MSC'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                       Period=c('2','2',
                               '2','2',
                               '2','2',
                               '2','2','2',
                               '2','2',
                               '2',
                               '2','2',
                               '2','2','2',
                               '2'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))

tmp_V10 <- merge(tmp_V10, df_trial_V10, by = "Patient", all.x = TRUE)

########################################################################


```

### Number of cells per CellType at 6m and 12m
```{r}

# ============================================================
# Supp_Figure_2 – Number of cells per CellType at 6m and 12m
# ============================================================
tmp1 <- tmp_V8
tmp1[tmp1$timePoint == "V8",  "timePoint"] <- "6m"

tmp2 <- tmp_V10
tmp2[tmp2$timePoint == "V10", "timePoint"] <- "12m"

tmp_6_12 <- rbind(tmp1, tmp2)

timePointCellType <- tmp_6_12 %>%
  dplyr::group_by(CellType, timePoint) %>%
  dplyr::summarise(samples = dplyr::n(), .groups = "drop")

timePointCellType$CellType  <- factor(timePointCellType$CellType, levels = cfg$all_cellType_order)
timePointCellType$timePoint <- factor(timePointCellType$timePoint, levels = c("6m", "12m"))

str_y <- "Number of cells"
o2 <- ggplot(timePointCellType) +
  aes(x = CellType, y = samples, fill = CellType) +
  geom_bar(stat = "identity", width = 0.28, alpha = 1, color = "black", linewidth = 0.2) +
  facet_wrap(~ timePoint, scales = "free_y", nrow = 3) +
  ylab(str_y) +
  theme_bw() +
  scale_fill_manual(values = cfg$colors_all) +
  theme(
    legend.position = "none",
    axis.title.y = element_text(size = 12),
    axis.text.y  = element_text(size = 12),
    axis.text.x  = element_text(angle = 90, size = 12, hjust = 0.95, vjust = 0.5),
    axis.title.x = element_blank(),
    aspect.ratio = 0.3
  )

ggsave(file.path(dirs$figures, "Supp_Figure_2.pdf"), o2, width = 12, height = 9)


```

### Lineage-level composition at 6 and 12 months by active treatment
```{r}

cluster_map <- data.frame(
  CellType = cfg$all_cellType_order,
  lineage = c('CD4_T','CD4_T','CD4_T','CD4_T','CD4_T',
              'CD8_T','CD8_T','CD8_T','CD8_T','CD8_T',
              'B','B','B','B','B','B',
              'NK','NK','NK','NK','NK',
              'DC','DC',
              'Mono','Mono'),
  stringsAsFactors = FALSE
)

lineage_colors <- c(
  CD4_T = "#1f77b4",
  CD8_T = "#9467bd",
  B     = "#d62728",
  NK    = "#ff7f0e",
  DC    = "khaki3",
  Mono  = "#2ca02c"
)

tmp_6_12_ann <- tmp_6_12 %>%
  dplyr::mutate(CellType = as.character(CellType)) %>%
  dplyr::left_join(cluster_map, by = "CellType")

lineage_counts <- tmp_6_12_ann %>%
  dplyr::group_by(Patient, timePoint, Treatment, lineage) %>%
  dplyr::summarise(n = dplyr::n(), .groups = "drop")

totals <- tmp_6_12_ann %>%
  dplyr::group_by(Patient, timePoint, Treatment) %>%
  dplyr::summarise(total_cells = dplyr::n(), .groups = "drop")

lineage_prop <- lineage_counts %>%
  dplyr::left_join(totals, by = c("Patient", "timePoint", "Treatment")) %>%
  dplyr::mutate(prop = n / total_cells) %>%
  dplyr::mutate(
    Patient = factor(Patient, levels = cfg$patientLevels_SMART),
    lineage = factor(lineage, levels = c("CD4_T","CD8_T","B","NK","DC","Mono"))
  )

o <- ggplot(lineage_prop, aes(x = Patient, y = prop, fill = lineage)) +
  geom_col() +
  facet_grid(timePoint ~ Treatment) +
  scale_fill_manual(values = lineage_colors) +
  labs(
    x = "Patient",
    y = "Relative abundance",
    fill = "Lineage",
    title = "Lineage-level composition at 6 and 12 months by active treatment"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    aspect.ratio = 0.5
  )

ggsave(file.path(dirs$figures, "Figure_3A.pdf"), o, width = 12, height = 7)


```



### UMAP only on V8/V10 subset + plot
```{r}


# 1) Subset FIRST (critical otherwise it will crash)
data_sub <- allCells[allCells$timePoint %in% c("V8", "V10"), ]

# 2) Ensure marker columns exist
missing_markers <- setdiff(col.names, colnames(data_sub))
if (length(missing_markers) > 0) {
  stop("Cannot compute UMAP: missing marker columns in data_sub: ",
       paste(missing_markers, collapse = ", "))
}

# 3) Optional subsampling (recommended if data_sub is large)
set.seed(params$seed)

keep_idx <- sample(seq_len(nrow(data_sub)), params$umap_max_cells)
data_umap <- data_sub[keep_idx, ]
  
# Save indices for full reproducibility/debugging
  
# 4) Compute UMAP on the chosen set only
X <- as.matrix(data_umap[,col.names])
u <- umap::umap(X)

data_umap$UMAP1 <- u$layout[, 1]
data_umap$UMAP2 <- u$layout[, 2]

# 5) Labels + timepoint names 
data_umap$labels <- factor(data_umap$CellType, levels = cfg$all_cellType_order)

data_umap$TimePoint <- data_umap$timePoint
data_umap[data_umap$TimePoint == "V8",  "TimePoint"] <- "6m"
data_umap[data_umap$TimePoint == "V10", "TimePoint"] <- "12m"
data_umap$TimePoint <- factor(data_umap$TimePoint, levels = c("6m", "12m"))

# 6) Plot (faithful styling)
umap_clust <- ggplot(data_umap, aes(x = UMAP1, y = UMAP2, color = labels)) +
  geom_point(size = 0.068) +
  coord_fixed(ratio = 1) +
  facet_wrap(~ TimePoint, nrow = 1) +
  theme_bw() +
  scale_color_manual(values = cfg$colors_all) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom",
    legend.background = element_rect()
  ) +
  guides(colour = guide_legend(override.aes = list(size = 3), title = ""))

ggsave(file.path(dirs$figures, "Supp_Figure_3.pdf"), umap_clust, width = 12, height = 6)

```

### cellFreqSummary (wide table) for the modelling section
```{r}

# ============================================================
# Generate cellFreqSummary (Patient x TimePoint wide frequencies)
# ============================================================

# Frequency per Patient × timePoint × CellType
freq_long <- allCells %>%
  dplyr::filter(Patient %in% cfg$patientLevels_SMART,
                timePoint %in% c("V4","V8","V10")) %>%
  dplyr::group_by(Patient, timePoint, CellType) %>%
  dplyr::summarise(samples = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Patient, timePoint) %>%
  dplyr::mutate(freq = samples / sum(samples)) %>%
  dplyr::ungroup()

# Wide format (one row per Patient×TimePoint, columns = all cell types)
cellFreqSummary <- freq_long %>%
  dplyr::select(Patient, TimePoint = timePoint, CellType, freq) %>%
  tidyr::pivot_wider(names_from = CellType, values_from = freq, values_fill = 0)

# Add Batch (unique batch per patient in that visit)
batch_map <- allCells %>%
  dplyr::filter(timePoint %in% c("V4","V8","V10")) %>%
  dplyr::group_by(Patient, timePoint) %>%
  dplyr::summarise(Batch = as.character(dplyr::first(Batch)), .groups = "drop") %>%
  dplyr::rename(TimePoint = timePoint)

cellFreqSummary <- cellFreqSummary %>%
  dplyr::left_join(batch_map, by = c("Patient","TimePoint"))

# Make sure all expected celltype columns exist (even if 0)
for (ct in cfg$all_cellType_order) {
  if (!ct %in% colnames(cellFreqSummary)) cellFreqSummary[[ct]] <- 0
}

# Reorder columns (optional but nice)
cellFreqSummary <- cellFreqSummary %>%
  dplyr::select(Patient, Batch, TimePoint, dplyr::all_of(cfg$all_cellType_order))

# Replace NAs with 0 
cellFreqSummary[is.na(cellFreqSummary)] <- 0


```


### Build modelling table (V8/V10) - CLR transform
```{r}
# Keep only V8/V10 for treatment modelling 
cellFreqSummary_6_12 <- cellFreqSummary %>%
  dplyr::filter(TimePoint %in% c("V8","V10"))

# Combine trial meta 
trial_meta <- dplyr::bind_rows(
  df_trial_V8  %>% dplyr::mutate(TimePoint = "V8"),
  df_trial_V10 %>% dplyr::mutate(TimePoint = "V10")
)

# Merge covariates onto frequencies (sample-level table)
cellFreqMeta <- cellFreqSummary_6_12 %>%
  dplyr::left_join(trial_meta, by = c("Patient","TimePoint"))

# Sanity check
if (any(is.na(cellFreqMeta$Treatment))) {
  stop("Missing clinical covariates after merge. Check Patient IDs and TimePoint labels.")
}

# Ensure all expected cell type columns exist
for (ct in cfg$all_cellType_order) {
  if (!ct %in% colnames(cellFreqMeta)) cellFreqMeta[[ct]] <- 0
}

# CLR transform with pseudocount
pseudocount <- 1e-6
props_mat <- as.matrix(cellFreqMeta[, cfg$all_cellType_order, drop = FALSE]) + pseudocount
props_mat <- props_mat / rowSums(props_mat)

log_mat <- log(props_mat)
gm <- rowMeans(log_mat)
clr_mat <- sweep(log_mat, 1, gm, "-")
colnames(clr_mat) <- paste0(cfg$all_cellType_order, "_clr")

cellFreqMeta <- dplyr::bind_cols(cellFreqMeta, as.data.frame(clr_mat))

# Make factors like the original model expects
cellFreqMeta <- cellFreqMeta %>%
  mutate(
    Patient  = factor(Patient),
    TimePoint = factor(TimePoint, levels = c("V8","V10")),
    Treatment = factor(Treatment),      # Control vs MSC
    Period   = factor(Period),
    Sequence = factor(Sequence),
    Sex      = factor(Sex),
    age_c    = scale(Age, center = TRUE, scale = FALSE)[,1]
  )

write.csv(cellFreqMeta, file.path(dirs$tables, "cellFreqMeta_CLR_V8_V10.csv"), row.names = FALSE)

```


### Mixed-effects modelling per cell type (lmer)
```{r}

clr_cols <- paste0(cfg$all_cellType_order, "_clr")
has_batch <- "Batch" %in% colnames(cellFreqMeta)

fit_one <- function(var_name) {

  rhs <- if (has_batch) {
    "Treatment + Period + Sequence + Batch + age_c + Sex + (1|Patient)"
  } else {
    "Treatment + Period + Sequence + age_c + Sex + (1|Patient)"
  }

  form <- as.formula(paste(var_name, "~", rhs))
  fit <- lmer(form, data = cellFreqMeta)
  coefs <- coef(summary(fit))

  trt_terms <- grep("^Treatment", rownames(coefs), value = TRUE)
  if (length(trt_terms) == 0) return(NULL)

  term <- if ("TreatmentMSC" %in% trt_terms) "TreatmentMSC" else trt_terms[1]

  data.frame(
    cell_type = var_name,
    term = term,
    estimate = coefs[term, "Estimate"],
    se = coefs[term, "Std. Error"],
    p_value = coefs[term, "Pr(>|t|)"],
    stringsAsFactors = FALSE
  )
}

results_comp <- map_dfr(clr_cols, fit_one)

results_comp <- results_comp %>%
  mutate(
    lower   = estimate - 1.96 * se,
    upper   = estimate + 1.96 * se,
    # back to original cell-type name (without _clr)
    CellType = sub("_clr$", "", cell_type),
    q_value = p.adjust(p_value, method = "BH")
  )

results_comp <- results_comp %>%
  mutate(
    Significant = q_value < 0.2,  # or p_value < 0.05, up to you
    Population  = CellType        # rename for plotting, if you want
  )

# Add lineage for colouring in plots
cluster_map <- data.frame(
  CellType = cfg$all_cellType_order,
  lineage = c('CD4_T','CD4_T','CD4_T','CD4_T','CD4_T',
              'CD8_T','CD8_T','CD8_T','CD8_T','CD8_T',
              'B','B','B','B','B','B',
              'NK','NK','NK','NK','NK',
              'DC','DC',
              'Mono','Mono'),
  stringsAsFactors = FALSE
)

results_comp <- results_comp %>%
  left_join(cluster_map, by = "CellType") %>%
  arrange(q_value)

write.csv(results_comp, file.path(dirs$tables, "Fig3_lmer_results_treatment_CLR.csv"), row.names = FALSE)
head(results_comp, 10)

# ============================================================
#  Forest plot of treatment effects (CLR)
# ============================================================

results_plot <- results_comp %>%
  mutate(CellType = factor(CellType, levels = rev(cfg$all_cellType_order)))

gg_forest <- ggplot(
  results_comp,
  aes(
    y = reorder(CellType, estimate),
    x = estimate,
    color = Significant
  )
) +
  geom_point(size = 3) +
  geom_errorbarh(
    aes(xmin = lower, xmax = upper),
    height = 0.2,
    alpha  = 0.7
  ) +
  geom_vline(xintercept = c(-0.25, 0.25), linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_vline(xintercept = 0, color = "gray60", linetype = "dashed") +
  scale_color_manual(
    values = c("TRUE" = "#D73027", "FALSE" = "gray60"),
    labels = c("FALSE" = "Not significant", "TRUE" = "Significant")
  ) +
  labs(
    title = "MSC vs Control effect on CLR-transformed abundance",
    x = "Effect size (CLR scale)",
    y = NULL,
    color = NULL
  ) +
  xlim(-1,1)+
  theme_minimal(base_size = 13)+
  theme(
    strip.background = element_rect(fill = "gray90", color = NA),
    strip.text = element_text(size = 13, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "bottom",
    aspect.ratio = 1,
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5)
  )

ggsave(file.path(dirs$figures, "Fig3_lmer_forest_treatment_CLR.pdf"), gg_forest, width = 10, height = 9)



o <- ggplot(results_comp, aes(x = estimate, y = -log10(p_value))) +
  geom_point(aes(color = Significant), size = 3, alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray60") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "gray60") +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "red", linewidth = 0.6) +
  geom_text_repel(
    #data = subset(results_df, adj.p.value < 0.05),
    aes(label = Population),
    size = 2.5,
    box.padding = 0.3,
    max.overlaps = 10
  ) +
  scale_color_manual(values = c("TRUE" = "#D73027", "FALSE" = "gray50")) +
  labs(
    title = "Volcano Plot of MSC Treatment",
    x = "Effect size (estimate) in CLR scale",
    y = paste0(expression(-log[10](p~value)),'- non Adjusted'),
    color = "Significant"
  ) +
  theme_minimal(base_size = 13)+
  theme(
    legend.position = "bottom",
    aspect.ratio = 1
  )

ggsave(file.path(dirs$figures, "lmer_volcano_treatment_CLR.pdf"), o, width = 10, height = 9)

```

### Diagnostics for top hits of modeling
```{r}

# ============================================================
# Diagnostics (top hits): residuals + QQ
# ============================================================

top_diag_n <- min(6, nrow(results_comp))

top_hits <- results_comp %>%
  filter(!is.na(q_value)) %>%
  slice(1:top_diag_n)

gg_residuals <- function(model, label) {
  df <- data.frame(Fitted = fitted(model), Residuals = resid(model))
  ggplot(df, aes(Fitted, Residuals)) +
    geom_point(alpha = 0.6) +
    geom_hline(yintercept = 0, color = "red") +
    theme_bw() +
    labs(title = paste("Residuals vs Fitted:", label))
}

gg_qq <- function(model, label) {
  df <- data.frame(Residuals = resid(model))
  ggplot(df, aes(sample = Residuals)) +
    stat_qq() + stat_qq_line() +
    theme_bw() +
    labs(title = paste("QQ plot:", label))
}

for (i in seq_len(nrow(top_hits))) {
  var_name <- top_hits$cell_type[i]
  label <- top_hits$CellType[i]
  
  rhs <- if ("Batch" %in% colnames(cellFreqMeta)) {
    "Treatment + Period + Sequence + Batch + age_c + Sex + (1|Patient)"
  } else {
    "Treatment + Period + Sequence + age_c + Sex + (1|Patient)"
  }
  
  fit <- lmer(as.formula(paste(var_name, "~", rhs)), data = cellFreqMeta)
  
  p1 <- gg_residuals(fit, label)
  p2 <- gg_qq(fit, label)
  
  ggsave(file.path(dirs$qc, paste0("diag_residuals_", label, ".pdf")), p1, width = 6, height = 5)
  ggsave(file.path(dirs$qc, paste0("diag_qq_", label, ".pdf")), p2, width = 6, height = 5)
}

```


### Extra code - Spaghetti plots on CLR scale for selected populations - EDIT if needed
```{r}


# Baseline SMART cohort
tmp0 <- tmp_V4[tmp_V4$Cohort == "SMART", ]
tmp0[tmp0$timePoint == "V4", "timePoint"] <- "BL"

# 
tmp <- rbind(tmp0, tmp1, tmp2)

tmp <- tmp %>%
  mutate(
    timePoint = factor(timePoint, levels = c("BL", "6m", "12m")),
    Treatment = factor(Treatment, levels = c("Baseline", "Control", "MSC"))
  )

# Populations you want to show
pops_of_interest <- c("CD4_CM", "CD4_TEMRA", "B_Mem", "NK_dim_CD16_low")


cell_counts <- tmp %>%
  group_by(Patient, timePoint, Treatment, CellType) %>%
  summarise(n = n(), .groups = "drop")

cell_counts_wide <- cell_counts %>%
  filter(CellType %in% pops_of_interest) %>%
  tidyr::pivot_wider(
    names_from  = CellType,
    values_from = n,
    values_fill = 0
  )

# extract count matrix for the 4 pops
mat <- as.matrix(cell_counts_wide[, pops_of_interest]) + 1e-6

clr_mat <- compositions::clr(mat)
colnames(clr_mat) <- paste0(pops_of_interest, "_clr")

cell_counts_clr <- cbind(
  cell_counts_wide[, c("Patient", "timePoint", "Treatment")],
  as.data.frame(clr_mat)
)

pop_counts_clr <- cell_counts_clr %>%
  tidyr::pivot_longer(
    cols      = ends_with("_clr"),
    names_to  = "CellType",
    values_to = "clr"
  ) %>%
  mutate(
    CellType = sub("_clr$", "", CellType),
    CellType = factor(CellType, levels = pops_of_interest),
    Patient  = factor(Patient)
  )

plot_violin_spaghetti_clr <- function(df, population_name) {
  ggplot(df, aes(x = timePoint, y = clr, fill = Treatment)) +
    geom_violin(trim = FALSE, alpha = 0.6) +
    geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8, color = "black") +
    geom_line(aes(group = Patient, color = Treatment), alpha = 0.4) +
    geom_point(
      aes(color = Treatment),
      size = 2,
      position = position_jitterdodge(jitter.width = 0.1)
    ) +
    labs(
      title = paste(population_name, ""),
      x     = "",
      y     = "CLR-transformed abundance"
    ) +
    scale_fill_manual(values = c(
      "Baseline" = "grey70",
      "Control"  = "royalblue3",
      "MSC"      = "khaki3"
    )) +
    scale_color_manual(values = c(
      "Baseline" = "grey40",
      "Control"  = "royalblue3",
      "MSC"      = "khaki3"
    )) +
    theme_minimal(base_size = 13) +
    theme(
      strip.background = element_rect(fill = "gray90", color = NA),
      strip.text       = element_text(size = 13, face = "bold"),
      axis.text.x      = element_text(size = 12, face = "bold"),
      axis.text.y      = element_text(size = 12),
      axis.title       = element_text(size = 14),
      legend.position  = "bottom",
      aspect.ratio     = 1,
      plot.title       = element_text(size = 15, face = "bold", hjust = 0.5)
    )
}

p_CD4_CM <- pop_counts_clr %>%
  filter(CellType == "CD4_CM") %>%
  plot_violin_spaghetti_clr("CD4_CM")

p_CD4_TEMRA <- pop_counts_clr %>%
  filter(CellType == "CD4_TEMRA") %>%
  plot_violin_spaghetti_clr("CD4_TEMRA")

p_B_Mem <- pop_counts_clr %>%
  filter(CellType == "B_Mem") %>%
  plot_violin_spaghetti_clr("B_Mem")

p_NK_dim <- pop_counts_clr %>%
  filter(CellType == "NK_dim_CD16_low") %>%
  plot_violin_spaghetti_clr("NK_dim_CD16_low")

fig3B_clr <- (p_CD4_CM | p_CD4_TEMRA) /
  (p_B_Mem  | p_NK_dim) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")


ggsave(file.path(dirs$figures, "Spaghetti_Figure_3C.pdf"), fig3B_clr, width = 12, height = 10)

```
