---
title: "SMART-MS Figure 4 – Functional scores in blood"
author: "Dimitrios Kleftogiannis"
output:
  html_document:
    toc: true
    toc_depth: 2
editor_options:
  chunk_output_type: inline
---
  
### Contact
  
Comments and bug reports are welcome, please email: Dimitrios Kleftogiannis (dimitrios.kleftogiannis@uib.no)

We are also interested to know about how you have used our source code, including any improvements that you have implemented.

You are free to modify, extend or distribute our source code, as long as our copyright notice remains unchanged and included in its entirety. 

### License

This code is licensed under the MIT License.

Copyright 2026, University of Bergen (UiB) and Neuro-SysMed, Norway

### Initiation and parameters setup - MAYBE EDITS ARE NEEDED FOR DIRS 
```{r}

suppressPackageStartupMessages({
  library(ggplot2)
  library(umap)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(circlize)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(lme4)
  library(lmerTest)
  library(patchwork)
  library(ggrepel)
})

#EDIT HERE IF NEEDED - SET YOUR WORKING DIR APPROPRIATELY
source("R/smartms_config.R")
source("R/io_zenodo.R")
#this is more elegant than before
cfg <- get_smartms_config()
# -----------------------------
# Parameters (EDIT THESE)
# -----------------------------
col.names <- c("CD16","CD4","CD14","CD19","CD3","CD235ab","CD11c","CD33",
                   "CD133","CD123","CD162","CD185","CD45RA","CD278","CD194","CD161",
                   "CD184","CD27","CD44","CD127","CD10","CD73","HLADR","CD146",
                   "CD117","CD8a","CD34","CD105","CD49d","CD20","CD25","CD66b",
                   "CD49f","CD45RO","CD90","CD45","CD195","CD38","CD196","CD135","CD56")
col.names.info <- c('type','type','type','type','type','state','type','type',
                        'state','type','state','state','type','state','state','state',
                        'state','state','state','state','state','state','type','state',
                        'state','type','type','state','state','type','state','type',
                        'state','type','state','type','state','type','state','state','type')

params <- list(
  # Direct download URL to Zenodo file that contains allCells_combined.RDa
  # (Use the "Download" link for the file on Zenodo)
  zenodo_url = "PASTE_ZENODO_DIRECT_DOWNLOAD_URL_HERE",

  #EDIT HERE IF NEEDED
  data_path = "PASTE_LOCAL_DOWNLOAD_FILE_HERE",
  # example -> data_path = '/Home/siv32/dkl081/Desktop/SMART/SMART_RAMMS_harmonised/allCells_combined.RDa',
  
  #EDIT HERE IF NEEDED
  out_base = "outputs/SMARTMS_publication",

  # Baseline definition for SMART-MS in your original notebook
  cohort = "SMART",
  timepoints = c("V4","V8","V10"),
  baseline_timePoint = "V4",
  baseline_label = "BL",

  # Reproducibility + runtime
  seed = 1,
  umap_max_cells = 50000,
  umap_markers = cfg$col.names, 
  
  # Modelling settings
  # cell type columns come from proportions (one per CellType)
  p_adjust_method = "BH",
  top_diag_n = 6  # number of top hits to show diagnostics
)

# -----------------------------
# Helpers
# -----------------------------
ensure_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE, showWarnings = FALSE)
}

download_if_missing <- function(url, dest_path) {
  if (file.exists(dest_path)) return(invisible(dest_path))
  ensure_dir(dirname(dest_path))
  message("Downloading dataset from Zenodo...")
  utils::download.file(url, destfile = dest_path, mode = "wb", quiet = FALSE)
  message("Saved to: ", dest_path)
  invisible(dest_path)
}

# Output directories for this figure
dirs <- list(
  figures = file.path(params$out_base, "figures", "fig4"),
  tables  = file.path(params$out_base, "tables",  "fig4"),
  qc      = file.path(params$out_base, "qc",      "fig4")
)
lapply(dirs, ensure_dir)

# Colors for time points (as in your original notebook)
color_TimePoints <- c("deepskyblue2", "magenta2", "indianred2")
names(color_TimePoints) <- params$timepoints

set.seed(params$seed)
```


### Load cleaned CyTOF data (Zenodo → local)
```{r}
download_if_missing(params$zenodo_url, params$data_path)

# This should load an object named: allCells
load(params$data_path)
stopifnot(exists("allCells"))

# Robust naming
if (!("CellType" %in% colnames(allCells)) && ("cellType" %in% colnames(allCells))) {
  allCells$CellType <- allCells$cellType
}

# Required columns 
required_cols <- c("Cohort","timePoint","Patient","CellType")
missing_cols <- setdiff(required_cols, colnames(allCells))
if (length(missing_cols) > 0) stop("Missing required columns in allCells: ", paste(missing_cols, collapse=", "))

# Keep SMART only
allCells <- allCells[allCells$Cohort == params$cohort, ]

# --- Subset by visit
tmp_V4  <- allCells[allCells$timePoint == "V4", ]
tmp_V8  <- allCells[allCells$timePoint == "V8", ]
tmp_V10 <- allCells[allCells$timePoint == "V10", ]

```


### Clinical design metadata 
```{r}

patientLevels_SMART <- cfg$patientLevels_SMART
################################################### V4
df_trial_V4 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline','Baseline',
                               'Baseline','Baseline',
                               'Baseline',
                               'Baseline','Baseline',
                               'Baseline','Baseline','Baseline',
                               'Baseline'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                       Period=c('0','0',
                               '0','0',
                               '0','0',
                               '0','0','0',
                               '0','0',
                               '0',
                               '0','0',
                               '0','0','0',
                               '0'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))
tmp_V4 <- merge(tmp_V4, df_trial_V4, by = "Patient", all.x = TRUE)

################################################### V8
df_trial_V8 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('Control','Control',
                               'MSC','MSC',
                               'Control','Control',
                               'MSC','MSC','MSC',
                               'Control','Control',
                               'MSC',
                               'Control','Control',
                               'MSC','MSC','MSC',
                               'Control'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                        Period=c('1','1',
                               '1','1',
                               '1','1',
                               '1','1','1',
                               '1','1',
                               '1',
                               '1','1',
                               '1','1','1',
                               '1'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))

tmp_V8 <- merge(tmp_V8, df_trial_V8, by = "Patient", all.x = TRUE)

################################################### V10
df_trial_V10 <- data.frame(Patient=patientLevels_SMART,
                         Treatment=c('MSC','MSC',
                               'Control','Control',
                               'MSC','MSC',
                               'Control','Control','Control',
                               'MSC','MSC',
                               'Control',
                               'MSC','MSC',
                               'Control','Control','Control',
                               'MSC'),
                       Sex=c('Male','Male',
                             'Male','Female',
                             'Female','Female',
                             'Female','Female',
                             'Male','Female',
                             'Male','Male',
                             'Female','Male',
                             'Male','Female',
                             'Female','Female'),
                       Period=c('2','2',
                               '2','2',
                               '2','2',
                               '2','2','2',
                               '2','2',
                               '2',
                               '2','2',
                               '2','2','2',
                               '2'),
                       Age=c(45,46,
                             53,42,
                             49,43,
                             51,46,
                             42,41,
                             52,47,
                             54,50,
                             45,40,
                             48,46),
                       Sequence=c('Control_first','Control_first',
                               'MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first','Control_first',
                               'MSC_first',
                               'Control_first','Control_first',
                               'MSC_first','MSC_first','MSC_first',
                               'Control_first'))

tmp_V10 <- merge(tmp_V10, df_trial_V10, by = "Patient", all.x = TRUE)

########################################################################

```

### Define functional scores from CyTOF markers
```{r}

# Each row = one marker's contribution to one score
score_def <- tribble(
  ~marker,   ~score,             ~role,
  # T-cell activation
  "HLA-DR",  "T_activation",     "pos",
  "CD38",    "T_activation",     "pos",
  "CD44",    "T_activation",     "pos",
  "CD25",    "T_activation",     "pos",
  "CD278",   "T_activation",     "pos",

  # T-cell memory / effector
  "CD45RO",  "T_mem_eff",        "pos",
  "CD27",    "T_mem_eff",        "pos",
  "CD127",   "T_mem_eff",        "pos",
  "CD45RA",  "T_mem_eff",        "neg",   # inverse

  # Th17 / inflammatory trafficking
  "CD196",   "Th17_homing",      "pos",
  "CD194",   "Th17_homing",      "pos",
  "CD195",   "Th17_homing",      "pos",
  "CD184",   "Th17_homing",      "pos",
  "CD161",   "Th17_homing",      "pos",
  "CD146",   "Th17_homing",      "pos",

  # CNS-homing T cells
  "CD49d",   "CNS_homing",       "pos",
  "CD184",   "CNS_homing",       "pos",
  "CD195",   "CNS_homing",       "pos",
  "CD49f",   "CNS_homing",       "pos",
  "CD162",   "CNS_homing",       "pos",

  # Treg-like (CD4+CD25+ context)
  "CD25",    "Treg_like",        "pos",
  "CD127",   "Treg_like",        "neg",   # inverse
  "CD73",    "Treg_like",        "pos",
  "HLA-DR",  "Treg_like",        "pos",

  # B-cell activation
  "CD38",    "B_activation",     "pos",
  "CD27",    "B_activation",     "pos",
  "CD73",    "B_activation",     "pos",
  "CD185",   "B_activation",     "pos",
  "HLA-DR",  "B_activation",     "pos",

  # B memory vs naive
  "CD27",    "B_mem_vs_naive",   "pos",
  "CD38",    "B_mem_vs_naive",   "pos",
  "CD45RO",  "B_mem_vs_naive",   "pos",
  "CD45RA",  "B_mem_vs_naive",   "neg",
  "CD10",    "B_mem_vs_naive",   "neg",

  # Myeloid APC activation
  "HLA-DR",  "Myeloid_APC",      "pos",
  "CD11c",   "Myeloid_APC",      "pos",
  "CD33",    "Myeloid_APC",      "pos",
  "CD38",    "Myeloid_APC",      "pos",
  "CD44",    "Myeloid_APC",      "pos",

  # Non-classical vs classical monocytes
  "CD14",    "Mono_nonclassical","neg",
  "CD16",    "Mono_nonclassical","pos",
  "HLA-DR",  "Mono_nonclassical","pos",

  # NK activation / maturation
  "CD56",    "NK_activation",    "pos",
  "CD16",    "NK_activation",    "pos",
  "CD161",   "NK_activation",    "pos",
  "CD38",    "NK_activation",    "pos",
  "CD44",    "NK_activation",    "pos"
)

markers <- sort(unique(score_def$marker))
scores  <- c("T_activation",
             "T_mem_eff",
             "Th17_homing",
             "CNS_homing",
             "Treg_like",
             "B_activation",
             "B_mem_vs_naive",
             "Myeloid_APC",
             "Mono_nonclassical",
             "NK_activation")

grid <- crossing(marker = markers, score = scores) %>%
  left_join(score_def, by = c("marker","score")) %>%
  mutate(
    role = replace_na(role, "none"),
    role = factor(role, levels = c("none","pos","neg")),
    score = factor(score, levels = scores),
    marker = factor(marker, levels = rev(markers))
  )

#### plot the score construction scheme
p_scores <- ggplot(grid, aes(x = score, y = marker, fill = role)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c(
      "none" = "grey95",
      "pos"  = "#D73027",  # red-ish for positive
      "neg"  = "#4575B4"   # blue-ish for inverse
    ),
    labels = c(
      "none" = "Not used",
      "pos"  = "Positive contribution",
      "neg"  = "Negative (inverse)"
    ),
    name = "Marker role"
  ) +
  labs(
    x = "Composite functional scores",
    y = "Markers",
    title = "Definition of composite functional scores"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x  = element_text(angle = 35, hjust = 1, vjust = 1),
    panel.grid   = element_blank(),
    plot.title   = element_text(face = "bold", hjust = 0.5),
    axis.title.y = element_text(margin = margin(r = 8)),
    legend.position = "bottom",aspect.ratio = 0.8
  )

ggsave(file.path(dirs$figures, "Figure_4A.pdf"), p_scores, width = 12, height = 9)

```

### Compute and visualise the functional scores at baseline
```{r}

# choose the markers needed for the scores
marker_cols <- c(
  "HLADR","CD38","CD44","CD25","CD278",          # activation
  "CD45RO","CD27","CD127","CD45RA",               # memory/eff
  "CD196","CD194","CD195","CD184","CD161","CD146",# Th17 / homing
  "CD49d","CD49f","CD162",                        # CNS-homing
  "CD73","CD11c","CD33","CD14","CD16",            # APC / mono
  "CD56","CD185","CD10"                           # NK / B mem-naive
)

tmp <- rbind(tmp_V4,tmp_V8,tmp_V10)

# baseline: Patient x CellType median marker expression
baseline_cluster <- tmp %>%
  group_by(Patient, timePoint,CellType) %>%
  summarise(
    across(all_of(marker_cols), median, na.rm = TRUE),
    .groups = "drop"
  )
#global z-score
baseline_z <- baseline_cluster %>%
  mutate(
    across(
      all_of(marker_cols),
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z"
    )
  )
#compute scores
baseline_scores <- baseline_z %>%
  mutate(
    # T-cell activation
    T_activation = (HLADR_z + CD38_z + CD44_z + CD25_z + CD278_z) / 5,

    # T-cell memory / effector differentiation
    T_mem_eff = (CD45RO_z + CD27_z + CD127_z - CD45RA_z) / 4,

    # Th17 / inflammatory trafficking
    Th17_homing = (CD196_z + CD194_z + CD195_z + CD184_z + CD161_z + CD146_z) / 6,

    # CNS-homing T cells
    CNS_homing = (CD49d_z + CD184_z + CD195_z + CD49f_z + CD162_z) / 5,

    # Treg-like (for CD4+CD25+ contexts)
    Treg_like = (CD25_z - CD127_z + CD73_z + HLADR_z) / 4,

    # B-cell activation
    B_activation = (CD38_z + CD27_z + CD73_z + CD185_z + HLADR_z) / 5,

    # B memory vs naive
    B_mem_vs_naive = (CD27_z + CD38_z + CD45RO_z - CD45RA_z - CD10_z) / 5,

    # Myeloid APC activation
    Myeloid_APC = (HLADR_z + CD11c_z + CD33_z + CD38_z + CD44_z) / 5,

    # Non-classical vs classical monocytes
    Mono_nonclassical = (-CD14_z + CD16_z + HLADR_z) / 3,

    # NK activation / maturation
    NK_activation = (CD56_z + CD16_z + CD161_z + CD38_z + CD44_z) / 5
  )

#subselect the score data frame
baseline_scores_short <- baseline_scores %>%
  select(
    Patient,
    timePoint,
    CellType,
    T_activation,
    T_mem_eff,
    Th17_homing,
    CNS_homing,
    Treg_like,
    B_activation,
    B_mem_vs_naive,
    Myeloid_APC,
    Mono_nonclassical,
    NK_activation
  )
#annotate cell types and lineages
cluster_annot <- data.frame(
  CellType = cfg$all_cellType_order,
  lineage  = c("CD4_T","CD4_T","CD4_T","CD4_T","CD4_T",
               "CD8_T","CD8_T","CD8_T","CD8_T","CD8_T",
               "B","B","B","B","B","B",
               "NK","NK","NK","NK","NK",
               "DC","DC",
               "Mono","Mono")
)

lineage_colors <- c(
  CD4_T = "#1f77b4",
  CD8_T = "#9467bd",
  B     =  "#d62728",
  NK    ="#ff7f0e",
  DC    = 'khaki3',
  Mono  = "#2ca02c"
)

score_cols <- c(
  "T_activation",
  "T_mem_eff",
  "Th17_homing",
  "CNS_homing",
  "Treg_like",
  "B_activation",
  "B_mem_vs_naive",
  "Myeloid_APC",
  "Mono_nonclassical",
  "NK_activation"
)

#subselect only baseline
baseline_scores_short <- baseline_scores_short[baseline_scores_short$timePoint=='V4',]

baseline_celltype_scores <- baseline_scores_short %>%
  group_by(CellType) %>%
  summarise(
    across(all_of(score_cols), median, na.rm = TRUE),
    .groups = "drop"
  )

baseline_celltype_scores <- baseline_celltype_scores %>%
  left_join(cluster_annot, by = "CellType")

lineage_order <- c("CD4_T","CD8_T","B","NK","DC","Mono")

baseline_celltype_scores <- baseline_celltype_scores %>%
  mutate(
    lineage  = factor(lineage, levels = lineage_order),
    CellType = factor(CellType, levels = cfg$all_cellType_order)  # your chosen order
  ) %>%
  arrange(lineage, CellType)

# matrix for heatmap (CellType × Scores)
mat_scores <- baseline_celltype_scores %>%
  select(all_of(score_cols)) %>%
  as.matrix()

rownames(mat_scores) <- baseline_celltype_scores$CellType

# row annotation: lineage per CellType
row_annot <- data.frame(
  lineage = baseline_celltype_scores$lineage
)
rownames(row_annot) <- baseline_celltype_scores$CellType

annot_colors <- list(
  lineage = lineage_colors
)

# colour palette for scores
score_palette <- colorRampPalette(c("#4575B4", "white", "#D73027"))(100)

out_file <- file.path(dirs$figures, "Functional_scores_baseline.pdf")

pheatmap::pheatmap(
  mat_scores,
  color            = score_palette,
  scale            = "none",         # scores are already z-based composites
  cluster_rows     = FALSE,
  cluster_cols     = FALSE,
  annotation_row   = row_annot,
  annotation_colors = annot_colors,
  border_color     = NA,
  fontsize_row     = 9,
  fontsize_col     = 10,
  main             = "Baseline functional scores per immune phenotype",
  gaps_row = c(5,10,16,21,23),
  gaps_col = c(5,7,9),
  filename =out_file
)

##### visualise scores for T cells - the code can be modified accoringly for other cell type subsets
baseline_long <- baseline_scores_short %>%
  select(Patient, CellType, all_of(score_cols)) %>%
  pivot_longer(
    cols      = all_of(score_cols),
    names_to  = "Score",
    values_to = "value"
  ) %>%
  left_join(cluster_annot, by = "CellType") %>%
  mutate(
    CellType = factor(CellType, levels = cfg$all_cellType_order),
    Score    = factor(Score, levels = score_cols)
  )

summarise_scores <- function(df) {
  df %>%
    group_by(Score, CellType) %>%
    summarise(
      median_val = median(value, na.rm = TRUE),
      se_val     = sd(value, na.rm = TRUE) / sqrt(n()),
      .groups    = "drop"
    )
}

plot_scores_block <- function(long_df, summary_df, title_text) {
  ggplot() +
    # median as filled circle
    geom_point(
      data = summary_df,
      aes(x = CellType, y = median_val, fill = CellType),
      shape = 21,
      size  = 2,
      color = "black"
    ) +
    # SE error bars
    geom_errorbar(
      data = summary_df,
      aes(
        x    = CellType,
        ymin = median_val - se_val,
        ymax = median_val + se_val
      ),
      width = 0.2
    ) +
    # individual patients as grey jittered points
    geom_point(
      data = long_df,
      aes(x = CellType, y = value, group = Patient),
      position = position_jitter(width = 0.1),
      alpha    = 0.3,
      size     = 1,
      color    = "grey48"
    ) +
    coord_flip() +
    facet_wrap(~ Score, scales = "free_x",nrow=1) +
    scale_fill_manual(values = cfg$colors_all) +
    labs(
      x     = "",
      y     = "Functional score (baseline)",
      title = title_text
    ) +
    theme_bw() +
    theme(
      panel.grid.major = element_line(size = 0.2),
      panel.grid.minor = element_blank(),
      legend.position  = "none",
      axis.text.x      = element_text(angle = 0, hjust = 0.5,size=6),
      axis.text.y      = element_text(size=6),
      strip.text       = element_text(face = "bold"),
      aspect.ratio = 1
    )
}

T_scores <- c("T_activation", "T_mem_eff", "Th17_homing", "CNS_homing", "Treg_like")
baseline_T <- baseline_long %>%
  filter(
    Score %in% T_scores,
    lineage %in% c("CD4_T", "CD8_T")
  )

summary_T <- summarise_scores(baseline_T)

p_T <- plot_scores_block(
  long_df    = baseline_T,
  summary_df = summary_T,
  title_text = "Baseline T-cell functional scores"
)

ggsave(file.path(dirs$figures, "Figure_4B.pdf"), p_T, width = 12, height = 9)

```


### Compute all functional scores after treatment and compare them between arms
```{r}
#use the combined data frame with all cells from BL, 6m and 12m
allCells <- tmp
rm(tmp) #save some space in memory

#redo the computation 
cluster_medians_all <- allCells %>%
  filter(timePoint %in% c("V4","V8","V10")) %>%     # BL, 6m, 12m
  group_by(Patient, timePoint, CellType, Batch, Age, Sex,
           Treatment, Period, Sequence) %>%         # assume these are already present
  summarise(
    across(all_of(marker_cols), median, na.rm = TRUE),
    .groups = "drop"
  )

cluster_z <- cluster_medians_all %>%
  mutate(
    across(
      all_of(marker_cols),
      ~ as.numeric(scale(.x, center = TRUE, scale = TRUE)),
      .names = "{.col}_z"
    )
  )

cluster_scores <- cluster_z %>%
  mutate(
    T_activation =
      (HLADR_z + CD38_z + CD44_z + CD25_z + CD278_z) / 5,

    T_mem_eff =
      (CD45RO_z + CD27_z + CD127_z - CD45RA_z) / 4,

    Th17_homing =
      (CD196_z + CD194_z + CD195_z + CD184_z + CD161_z + CD146_z) / 6,

    CNS_homing =
      (CD49d_z + CD184_z + CD195_z + CD49f_z + CD162_z) / 5,

    Treg_like =
      (CD25_z - CD127_z + CD73_z + HLADR_z) / 4,

    B_activation =
      (CD38_z + CD27_z + CD73_z + CD185_z + HLADR_z) / 5,

    B_mem_vs_naive =
      (CD27_z + CD38_z + CD45RO_z - CD45RA_z - CD10_z) / 5,

    Myeloid_APC =
      (HLADR_z + CD11c_z + CD33_z + CD38_z + CD44_z) / 5,

    Mono_nonclassical =
      (-CD14_z + CD16_z + HLADR_z) / 3,

    NK_activation =
      (CD56_z + CD16_z + CD161_z + CD38_z + CD44_z) / 5
  )

#generate the data frame with all info we need for modeling
analysis_scores_df <- cluster_scores %>%
  filter(timePoint %in% c("V8","V10"),
         Treatment %in% c("Control","MSC")) %>%
  mutate(
    Treatment = factor(Treatment, levels = c("Control","MSC")),
    Period    = factor(Period, levels = c("1","2")),
    Sequence  = factor(Sequence, levels = c("Control_first","MSC_first")),
    Sex       = factor(Sex),
    age_c     = Age - mean(Age, na.rm = TRUE)
  )

analysis_scores_df_short <- analysis_scores_df %>%
  select(
    Patient,
    CellType,
    Batch,
    age_c,
    Sex,
    Period,
    timePoint,
    Sequence,
    Treatment,
    T_activation,
    T_mem_eff,
    Th17_homing,
    CNS_homing,
    Treg_like,
    B_activation,
    B_mem_vs_naive,
    Myeloid_APC,
    Mono_nonclassical,
    NK_activation
  )

analysis_scores_df_short <- analysis_scores_df_short %>%
  left_join(cluster_annot, by = "CellType")

```


### Longitudinal modelling of functional scores 
We use similar setup with the cell composition analysis
```{r}

# main function with formula used
fit_one_score_celltype <- function(df, score_name, celltype_name) {
  # subset to this CellType
  df_sub <- df[df$CellType == celltype_name, ]

  # Guard: not enough data or only one treatment present
  if (nrow(df_sub) < 6 || length(unique(df_sub$Treatment)) < 2) {
    return(NULL)
  }

  # build the formula: score ~ Treatment + Period + Sequence + Batch + age_c + Sex + (1|Patient)
  form <- as.formula(
    paste0(score_name, " ~ Treatment + Period + Sequence + Batch + age_c + Sex + (1 | Patient)")
  )

  fit <- lmer(form, data = df_sub)
  coefs <- coef(summary(fit))

  # robustly check for TreatmentMSC row
  if (!"TreatmentMSC" %in% rownames(coefs)) {
    return(NULL)
  }

  data.frame(
    CellType = celltype_name,
    Score    = score_name,
    estimate = coefs["TreatmentMSC", "Estimate"],
    se       = coefs["TreatmentMSC", "Std. Error"],
    p_value  = coefs["TreatmentMSC", "Pr(>|t|)"],
    stringsAsFactors = FALSE
  )
}

res_list <- list()
k <- 1
for (sc in score_cols) {
  for (ct in cfg$all_cellType_order) {
    out <- fit_one_score_celltype(
      df           = analysis_scores_df,
      score_name   = sc,
      celltype_name = ct
    )
    if (!is.null(out)) {
      res_list[[k]] <- out
      k <- k + 1
    }
  }
}
results_scores <- do.call(rbind, res_list)
results_scores <- results_scores %>%
  mutate(
    lower   = estimate - 1.96 * se,
    upper   = estimate + 1.96 * se,
    q_value = p.adjust(p_value, method = "BH")
  )
results_scores <- results_scores %>% left_join(cluster_annot, by = "CellType")


```


### Visualisation - Heatmap with effect sizes
```{r}

# Use the same score order as in your baseline heatmap
score_order <- colnames(mat_scores)   # assuming mat_scores already in correct col order

# Use the same cell-type order as in row_annot / mat_scores
celltype_order <- rownames(mat_scores)

effects_wide <- results_scores %>%
  mutate(
    Score    = factor(Score, levels = score_order),
    CellType = factor(CellType, levels = celltype_order)
  ) %>%
  arrange(CellType, Score) %>%
  select(CellType, Score, estimate) %>%
  tidyr::pivot_wider(
    names_from  = Score,
    values_from = estimate
  )

mat_effects <- as.matrix(effects_wide[ , score_order, drop = FALSE])
rownames(mat_effects) <- effects_wide$CellType

effect_palette <- colorRampPalette(c("#f1a340", "#f7f7f7", "#998ec3"))(100)

max_abs <- max(abs(mat_effects), na.rm = TRUE)
breaks_effects <- seq(-max_abs, max_abs, length.out = 100)

num_mat <- ifelse(results_scores$ p_value < 0.05, "*", "")
num_wide <- results_scores %>%
  mutate(
    Score    = factor(Score, levels = score_order),
    CellType = factor(CellType, levels = celltype_order),
    star     = ifelse(p_value < 0.05, "*", "")
  ) %>%
  select(CellType, Score, star) %>%
  pivot_wider(names_from = Score, values_from = star)

num_mat <- as.matrix(num_wide[ , score_order, drop = FALSE])
rownames(num_mat) <- num_wide$CellType

out_file <- file.path(dirs$figures, "Figure_4C.pdf")

pheatmap::pheatmap(
  mat_effects,
  color             = effect_palette,      # <-- new palette
  breaks            = breaks_effects,
  scale             = "none",
  cluster_rows      = FALSE,
  cluster_cols      = FALSE,
  annotation_row    = row_annot,
  annotation_colors = annot_colors,
  border_color      = NA,
  fontsize_row      = 9,
  fontsize_col      = 10,
  main              = "MSC vs Control effect on functional scores",
  gaps_row          = c(5,10,16,21,23),
  gaps_col          = c(5,7,9),
  display_numbers   = num_mat,
  filename          = out_file
)

```
